% Word count aim: 4000-4500
% current: 3293
\chapter{Background}

\section{Computational Problems}

\subsection{Problem Formulations}
% Define problem instances and problem classes.
% Distinguish between decision, search, and optimization problems.
% Clarify deterministic vs. stochastic problem formulations.

A \textit{computational problem} is a formal specification of a task defined over a set of
\textit{problem instances}. Each instance encodes the input data relevant to a particular
occurrence of the problem. The \textit{problem class} describes the collection of all
such problem instances together with the corresponding output requirements.~\cite{HOOS200513}
 
Problems are broadly categorized by the nature of their required output. 
\begin{itemize}
    \item A \textit{decision problem} demands a binary answer, whether a solution exists for a problem instance. 
    \item A \textit{search problem} asks for the construction of an explicit solution
    satisfying a specified property, such as a feasible assignment or a path in a graph, or conclude no solution exists.
    \item An \textit{optimization problem} further requires that the returned solution be optimal (or near-optimal) with 
    respect to some given objective function.
\end{itemize}
The decision and search problem variants are closely related because algorithms solving the 
search variant can always be used to solve the decision variant.~\cite{HOOS200513}

A further distinction concerns the role of randomness in the problem formulation. In a
\textit{deterministic} formulation, both the instance and every algorithm operating on it are fully specified without any probabilistic component. In a \textit{stochastic}
formulation, the instance itself, the objective evaluations, or the algorithm may involve randomness. Noisy objective functions, where repeated evaluation of the same candidate
solution may return different values, are a canonical example and arise naturally in
simulation-based and real-world settings.

\subsection{Algorithms}
\subsubsection{Algorithm Properties}
\paragraph{Deterministic and Randomized Algorithms. }
% Definitions and differences.

\paragraph{Adaptive vs. Non-Adaptive Algorithms}
% Sequential decision processes.
% Role of feedback.

\subsubsection{Oracle-Based Computation}
% Definition of oracle access.
% Query-response interaction model.

An \textit{oracle} is an abstract computational device that answers queries about a
problem instance. In optimization, the most common oracle is the \textit{function
evaluation oracle}, which, given a query point $x \in \mathcal{X}$, returns the
objective value $f(x)$ (or a noisy estimate thereof). The algorithm interacts with the
oracle in a \textit{query-response} model: it submits a query and receives a response
before deciding on the next query. The total number of such interactions constitutes the
\textit{query complexity} of the algorithm, formalized in
Section~\ref{subsubsec:infocomplexity}~\cite{nocedal2006numerical, droste2006blackbox}.



\subsection{Search Space}
\subsubsection{Structure of the Search Space}
% Finite vs. infinite spaces.
% Discrete vs. continuous domains.
% Combinatorial vs. geometric domains.

The \textit{search space} $\mathcal{X}$ is the set of all candidate solutions considered
by an algorithm. Search spaces may be \textit{finite} or \textit{infinite}, and
\textit{discrete} or \textit{continuous}. 
 
Discrete combinatorial search spaces, in particular the Boolean hypercube $\{0,1\}^n$, are central to the theoretical analysis of evolutionary algorithms (\acrshort{ea}) and other randomized search
heuristics~\cite{droste2002analysis,HOOS200561} used in theoretical \acrshort{bbo} analysis. 
Continuous spaces $\mathbb{R}^d$ arise in numerical optimization~\cite{nocedal2006numerical}, and mixed spaces containing both discrete and continuous variables occur in practice but are considerably harder to
analyze~\cite{audet2017derivative}.
 
The structure of the search space, and in particular how neighboring solutions relate to
one another, fundamentally influences the difficulty of optimization. This is encoded by
a \textit{neighborhood function} $\mathcal{N} : \mathcal{X} \to 2^{\mathcal{X}}$,
assigning to each candidate solution the set of solutions reachable in a single step!CITE: Hoos and Stützle - 2005 - INTRODUCTION!.
For the Boolean hypercube, the standard neighborhood consists of all strings at Hamming
distance one !CITE: a resource using hamming distance!.

\begin{remark}[Problem domain of interest]
This project is concerned with \emph{Pseudo-Boolean, Combinatorial Optimization problems}, with solution space $\mathcal{X} \subseteq \{0,1\}^n$. A candidate solution can be represented as a \emph{bitstring}: a sequence of alphabets $a \in \{0,1\}$. The objective functions considered for such problems are of the form $f:\mathcal{X} \to \mathbb{R}_0^+$, where $\mathbb{R}_0^+$ denotes the non-negative reals. Without loss of generality, an objective function with co-domain $\mathcal{T} \subseteq \mathbb{R}$ can be mapped to this convention via the affine shift
\[
    g: \mathbb{R} \to \mathbb{R}_0^+, \qquad g(x) = x - \min_{y \in \mathcal{X}} f(y),
\]
which is well-defined whenever $f$ attains its minimum on $\mathcal{X}$, and preserves the ordering of all objective values by construction.
\end{remark}


\subsubsection{Constraints and Feasibility}
% Feasible region definition.
% Constrained vs. unconstrained problems.

In a \textit{constrained} optimization problem, the \textit{feasible region}
$\mathcal{F} \subseteq \mathcal{X}$ is defined by one or more constraints that a valid
solution must satisfy~\cite{nocedal2006numerical, bertsekas2009convex}. 
An \textit{unconstrained} problem places no restriction beyond
membership in $\mathcal{X}$. Constraints may be \textit{hard} (a solution violating them
is inadmissible) or \textit{soft} (violations are penalized in the objective). The
boundary of the feasible region can significantly complicate optimization, particularly
for local search methods that may generate infeasible intermediate solutions.

\subsection{Objective Functions and Landscape Properties}

The geometric structure of the objective function determines both which algorithms are
applicable and what convergence guarantees are attainable. A central organizing
principle is that access to higher-order analytical information about $f$, i.e., gradients, 
curvature, or structural properties such as convexity, generally enables stronger
theoretical guarantees and more efficient algorithms; restricting this access,
as in the black-box setting, correspondingly limits what can be guaranteed.
 
\subsubsection{Zeroth-, First-, and Second-Order Methods}
% Keep existing content; add forward reference to §2.2 (BBO = zeroth-order world)
 
Optimization algorithms can be classified by the order of analytical information they
require from $f$. This \textit{information hierarchy} has three tiers.
 
\textit{Zeroth-order} methods use only evaluations of $f(x)$, with no gradient or
curvature information. The algorithm observes only the output of the function evaluation
oracle at each queried point and must infer search directions from comparisons of
objective values alone. This exactly describes the black-box model formalized in
Section~\ref{sec:bbo}: when the internal structure of $f$ is inaccessible or
analytically intractable, zeroth-order methods are the only option. Their weaker
information access is the fundamental reason black-box optimization is harder in
general than white-box optimization, and it is the regime in which all algorithms
studied in this project operate.
 
\textit{First-order} methods additionally access the gradient $\nabla f(x) \in
\mathbb{R}^d$, which encodes the direction of steepest ascent at $x$. For a
differentiable function $f : \mathbb{R}^d \to \mathbb{R}$, gradient-based methods
exploit this to define descent directions and form the backbone of classical continuous
optimization~\cite{nocedal2006numerical}. When $f$ is convex but non-differentiable,
the gradient generalizes to the \textit{subdifferential} $\partial f(x)$, the set of
all \textit{subgradients} $g$ satisfying
\[
  f(y) \geq f(x) + g^\top(y - x) \quad \forall\, y \in \mathbb{R}^d.
\]
Subgradient methods replace the gradient with any element of $\partial f(x)$, recovering
convergence under convexity alone, albeit at rates slower than those attainable with
smooth gradients~\cite{bertsekas2009convex}.
 
\textit{Second-order} methods further access the Hessian $\nabla^2 f(x) \in
\mathbb{R}^{d \times d}$, which captures local curvature. Methods that utilize or
approximate the Hessian, such as Newton and quasi-Newton schemes, can achieve
superlinear or quadratic convergence near a solution~\cite{nocedal2006numerical}, at the
cost of maintaining a $d \times d$ matrix.
 
Together, these three tiers define distinct levels of oracle access, and their
availability directly governs the choice and efficiency of applicable algorithms. The
move from second-order to zeroth-order corresponds to progressively weaker assumptions
about what the oracle reveals, culminating in the black-box model where only function
values are observable.
 
\subsubsection{Landscape Properties}
 
The following geometric properties of $f$ characterize which tier of analytical
information is both available and useful.
 
\paragraph{Convexity.}
A function $f : \mathbb{R}^d \to \mathbb{R}$ is \textit{convex} if for all
$x, y \in \mathbb{R}^d$ and $\lambda \in [0,1]$~\cite{bertsekas2009convex},
\[
  f\!\left(\lambda x + (1-\lambda)y\right) \leq \lambda f(x) + (1-\lambda)f(y).
\]
Convexity guarantees that every local minimum is global, rendering the landscape
tractable for first-order methods. \textit{Strong convexity} additionally requires
\[
  f(y) \geq f(x) + \nabla f(x)^\top(y - x) + \tfrac{\mu}{2}\|y - x\|^2
\]
for some $\mu > 0$, implying a unique global minimum and linear convergence rates for
gradient descent~!CITE: Nesterov 2004!.
 
\paragraph{Smoothness and Lipschitz Continuity.}
% TODO
 
\paragraph{Ruggedness and Multimodality.}
% - Multiple local optima, deceptive gradients, epistatic interactions
% - Motivates heuristic / stochastic methods: no first-order structure to exploit
% - Forward reference to NK-landscapes as a tunable model of ruggedness

\subsection{Complexity Measures}
\subsubsection{Classical Complexity: Time and Space}
% Standard asymptotic measures.

\subsubsection{Information Complexity: Query and Oracle Complexity}
\label{subsubsec:infocomplexity}
% Number of oracle calls as complexity measure.

\subsubsection{Expected vs. Worst-Case Analysis}
% Runtime notions in randomized algorithms.


% \subsection{Probabilistic Tools for Algorithm Analysis}
% \subsubsection{Expectation and Variance}
% \subsubsection{Concentration Inequalities}
% \paragraph{Markov's Inequality. }
% \paragraph{Chebyshev's Inequality. }
% \paragraph{Chernoff Bounds. }

\newpage

\section{Black-Box Optimization}
\label{sec:bbo}

\subsection{The Black-Box Model}
% Formal oracle model.
% Zeroth-order information.
% Information revealed per query.
% - What structural knowledge white-box methods exploit
% - Gradient availability as the dividing line
% - Computational trade-offs: information richness vs. access cost

In the \textit{black-box optimization} model, the algorithm has no access to the
analytical form of the objective function. It interacts with the objective exclusively
through an oracle that, upon receiving a query $x \in \mathcal{X}$, returns $f(x)$ (or
a noisy estimate thereof)~\cite{droste2006blackbox,doerr2020theory,lengler2019drift}. 
This constitutes \textit{zeroth-order} information: no gradient or higher-order information is available.
The black-box model is appropriate whenever the objective arises from a simulation,
physical experiment, or any process whose internal structure is inaccessible or too
complex to exploit analytically.

In contrast, in \textit{white-box optimization}, information about the problem instance is make available to 
the algorithm. This enables algorithms usage of some computational tools such as gradients, supporting
optimization convergence.

\subsubsection{Variants of Black-Box Models}
% Unrestricted model.
% Ranking-based model.
% Memory-restricted model.


\subsection{Black-Box Complexity}
\label{subsec:bb-complexity}
% - Definition: minimum number of queries any algorithm requires in the worst case
%   to solve a given problem class  [droste2006blackbox]
% - Yao's Minimax Principle: lower bounds via distributions over instances
% - Upper bounds (algorithmic) and lower bounds (information-theoretic)
% - Comparison-based lower bounds (sorting analogy)
% - What these results say about the hardness of combinatorial BBO

\subsection{Purpose of Randomized Search Heuristics}
% Explain why, given the BBO setting, one turns to general-purpose
% stochastic methods rather than problem-specific exact algorithms.
%
% - No-Free-Lunch Theorems: statement and implications for algorithm comparison
%   [wolpert1997no or similar]
%   No universally superior algorithm exists across all problem instances;
%   the choice of algorithm embeds implicit assumptions about problem structure.
% - Practical consequences:
%     When the function structure is unknown, heuristic methods are natural
%     When exact methods are computationally infeasible (NP-hard instances),
%     anytime heuristics offer the best available solution within a budget
%     Stochasticity enables escape from local optima that deterministic
%     local search cannot avoid
% - Taxonomy of derivative-free / black-box methods:
%     Exact DFO (Nelder-Mead, pattern search, model-based DFO)
%     Bayesian Optimization (Gaussian process surrogate + acquisition)
%     Evolutionary / population-based heuristics
%     Single-solution local search heuristics
%     Simulated Annealing and other acceptance-criterion methods
% - Scope of this project: randomized single- and population-based heuristics
%   on combinatorial / pseudo-Boolean domains
 
\subsubsection{No-Free-Lunch Theorems}
 
\subsubsection{Taxonomy of Derivative-Free Methods}
% Brief map of the full space; help the reader locate RSHs within it


\subsection{Randomized Search Heuristic Algorithms}


\subsubsection[\texorpdfstring{Evolutionary Algorithms (EA)}{Evolutionary Algorithms (EA)}]{Evolutionary Algorithms (\acrshort{ea})}
\label{subsubsec:ea}

Evolutionary algorithms are a broad family of population-based, randomized search
heuristics inspired by biological evolution. In discrete optimization, the canonical
paradigm maintains one or more \emph{candidate solutions} (bit strings) and iterates
two operators: \emph{mutation}, which perturbs a solution stochastically, and
\emph{selection}, which retains solutions according to their objective
value~\cite{droste2002analysis,doerr2020theory}. The simplest instance of this
paradigm, the $(1+1)$~EA, serves as the theoretical baseline throughout this project.

\paragraph{The $(1+1)$~EA.}
\label{par:oneone-ea}
The $(1+1)$~EA maintains a single current solution $x \in \{0,1\}^n$. In each
iteration it produces one offspring $y$ by \emph{standard bit mutation}: every bit is
flipped independently with probability $p$. The offspring replaces the parent if and
only if $f(y) \geq f(x)$ (\emph{elitist selection}). With the standard choice $p =
1/n$ the expected number of flipped bits is exactly one. The resulting process is a
homogeneous Markov chain on $\{0,1\}^n$ with a non-decreasing fitness
sequence~\cite{droste2002analysis,witt2013tight}.

\begin{algorithm}[H]
\caption{The $(1+1)$~EA with mutation probability $p \in (0,1)$}
\label{alg:oneone-ea}
\begin{algorithmic}[1]
  \State Sample $x \in \{0,1\}^n$ uniformly at random; evaluate $f(x)$
  \Repeat
    \State $y \gets x$
    \For{$i = 1, \ldots, n$}
      \State flip $y_i$ with probability $p$, independently
    \EndFor
    \If{$f(y) \geq f(x)$} $x \gets y$ \EndIf
  \Until{termination criterion met}
\end{algorithmic}
\end{algorithm}

\noindent The classical runtime results for $p = 1/n$ are
\[
  \mathbb{E}[T_{\textsc{OneMax}}] = e\,n\ln n + O(n)
  \qquad \text{and} \qquad
  \mathbb{E}[T_{\textsc{LeadingOnes}}] = \tfrac{e-1}{2}\,n^2 + o(n^2),
\]
and the $\Theta(n\log n)$ bound extends to \emph{all} linear pseudo-Boolean functions
with non-zero weights~\cite{witt2013tight}. On general landscapes the algorithm may require
exponential time: even simple quadratic and unimodal functions can trap it
indefinitely without restarts or adaptive mutation~\cite{droste2002analysis}. Advanced variants and
implementation-aware refinements are discussed in
Appendix~\ref{app:ea}.


\subsubsection{Simulated Annealing}


\paragraph{Cooling Schedules}



\subsubsection[\texorpdfstring{Randomized Local Search (RLS)}{Randomized Local Search (RLS)}]{\acrfull{rls}}




\paragraph[\texorpdfstring{Stochastic Hill-Climber - Greedy RLS}{Stochastic Hill-Climber - Greedy RLS}]{Stochastic Hill-Climber - Greedy \acrshort{rls}}



\paragraph[\texorpdfstring{$\epsilon$-greedy RLS}{epsilon-greedy RLS}]{$\epsilon-greedy$ \acrshort{rls}}



\paragraph[\texorpdfstring{Decaying $\epsilon-greedy$ RLS}{Decaying epsilon-greedy RLS}]{Decaying $\epsilon-greedy$ \acrshort{rls}}



\subsection{Few Pseudo-Boolean Problems}

The \textit{pseudo-Boolean} setting considers functions $f : \{0,1\}^n \to \mathbb{R}$
defined on binary strings of length $n$. Several benchmark functions are widely used to
study and compare the behavior of randomized search heuristics, each encoding a
different type of difficulty.~\cite{lengler2019drift,droste2006blackbox}

\subsubsection{OneMax}

The \textsc{OneMax} function counts the number of one-bits in a bit string:
\[
  \textsc{OneMax}(x) = \sum_{i=1}^{n} x_i, \qquad x \in \{0,1\}^n.
\]
The unique global maximum is $x^{*} = 1^n$ with $\textsc{OneMax}(1^n) = n$.
\textsc{OneMax} is the canonical baseline benchmark: the landscape is unimodal and every
improving move increases the function value by exactly one, providing a clear gradient
for local search methods. The expected runtime of the $(1+1)$~\acrfull{ea} on \textsc{OneMax} is
$\Theta(n \log n)$. !CITE: Droste, Jansen, Wegener (2002)!

\subsubsection{LeadingOnes}

The \textsc{LeadingOnes} function counts the length of the maximal leading prefix of
ones:
\[
  \textsc{LeadingOnes}(x) = \sum_{i=1}^{n} \prod_{j=1}^{i} x_j.
\]
Equivalently, $\textsc{LeadingOnes}(x)$ equals the index of the first zero bit minus
one, or $n$ if no zero exists. The unique global optimum is again $1^n$. Because
progress requires the leading bit to be set before any subsequent bit can contribute,
\textsc{LeadingOnes} is significantly harder than \textsc{OneMax}: the expected runtime
of the $(1+1)$~\acrshort{ea} is $\Theta(n^2)$. !CITE: Droste, Jansen, Wegener (2002)!


\subsubsection{Jump}

The \textsc{Jump} function with parameter $k \in \{1, \ldots, n\}$ coincides with
\textsc{OneMax} on bit strings with at most $n - k$ ones and on the global optimum
$1^n$, but introduces a region of reduced fitness for strings with between $n-k+1$ and
$n-1$ ones. This creates a ``gap'' of width $k$ that algorithms relying on single-bit
improvements cannot cross: the set of strings with exactly $n-k$ ones constitutes a
local optimum of standard local search, and reaching the global optimum $1^n$ requires
flipping all $k$ remaining zero bits simultaneously. For the $(1+1)$~\acrshort{ea} with standard
bit mutation at rate $1/n$, the probability of a specific $k$-bit flip is
$\Theta(n^{-k})$, giving an expected runtime of $\Theta(n^k)$ on $\textsc{Jump}_k$.

Formally, \textsc{Jump}$_{k,n} \colon \{0,1\}^n \to \mathbb{R}$ is defined as
\[
  \textsc{Jump}_{k,n}(x) \;=\;
  \begin{cases}
    k + \displaystyle\sum_{i=1}^{n} x_i & \text{if } \displaystyle\sum_{i=1}^{n} x_i \leq n - k
    \text{ or } \displaystyle\sum_{i=1}^{n} x_i = n, \\[8pt]
    n - \displaystyle\sum_{i=1}^{n} x_i & \text{otherwise.}
  \end{cases}
\]
Droste, Jansen, and Wegener~\cite{droste2002analysis} proved that the expected running
time of the $(1+1)$~\acrshort{ea} on $\textsc{Jump}_{k,n}$ is $\Theta(n^k + n\log n)$ for
$k \in \{1, 2, \ldots, n\}$, which simplifies to $\Theta(n^k)$ for $k \geq 2$.
\subsubsection{BinaryValue[BinVal]}
The \textsc{BinaryValue} (\textsc{BinVal}) function interprets a bit string as a binary
integer by weighting bits exponentially by position:
\[
  \textsc{BinVal}(x) = \sum_{i=1}^{n} 2^{i-1}\, x_i.
\]
The global optimum is $1^n$ with value $2^n - 1$. The exponential weighting means that
the most significant bit contributes more than the combined value of all less significant
bits, making \textsc{BinVal} highly structured.

\subsubsection{Trap}
\subsubsection{Plateau}
\subsubsection{Hierarchical If-Only-If}


\subsubsection{MaxSAT}
\textit{Maximum Satisfiability} (MaxSAT) is the optimization variant of the Boolean
satisfiability problem (SAT). Given a propositional formula in conjunctive normal form (CNF) !TODO: Cite or explain CNF!
with $m$ clauses over $n$ Boolean variables, the goal is to find a truth assignment that
maximizes the number of satisfied clauses. MaxSAT is NP-hard in general~\cite{hutchison_search_2004}
and serves as an important benchmark for combinatorial optimization algorithms.

\subsubsection{Tunable Landscapes: NK-Landscapes}

\textit{NK-landscapes}~\cite{kauffman1987towards} are a parametric family of pseudo-Boolean
functions controlled by two integers: $N$ (the number of variables) and $K$ (the number
of epistatic interactions per variable, $0 \leq K \leq N-1$). The fitness of a string
$x \in \{0,1\}^N$ is defined as
\[
  f(x) = \frac{1}{N} \sum_{i=1}^{N} f_i\!\left(x_i,\, x_{\pi(i,1)},
      \ldots,\, x_{\pi(i,K)}\right),
\]
where each local contribution $f_i : \{0,1\}^{K+1} \to [0,1]$ is drawn independently
and uniformly at random, and $\pi(i, \cdot)$ denotes the $K$ epistatic neighbours of
variable $i$.

When $K = 0$ the contributions are independent, producing a smooth, additively separable
landscape analogous to \textsc{OneMax}. As $K$ increases toward $N-1$, epistatic
interactions intensify and the landscape becomes increasingly rugged, with the number of
local optima growing in the limit. NK-landscapes thus provide a tunable model of
ruggedness and are widely used to study algorithmic scalability and robustness across a
spectrum of problem difficulties.

\subsection{Traditional Performance Measures in BBO}
\label{subsec:bbo-perf}
% - Expected runtime (hitting time to reach the optimum or a quality threshold)
%   as the standard measure in EA theory  [droste2002analysis; doerr2020theory]
% - Fixed-budget analysis: quality of best-found solution after T queries
%   [Jansen & Zarges 2014 or similar]
% - Fixed-target analysis: minimum budget to reach a target quality with
%   probability >= 1-delta  [!CITE!]
% - Limitations of these measures:
%     Binary in nature (found / not found)
%     Ignore the trajectory of quality improvement
%     Do not penalise "wasteful" queries on already-known-poor regions
% - This gap motivates regret as a more informative measure
The theoretical analysis of \acrshort{bbo} algorithms has developed several complementary
frameworks for quantifying performance. Each framework fixes one dimension of the
problem, either the evaluation budget or the target solution quality, and asks what
can be guaranteed along the other. Together they form the standard vocabulary for
comparing randomized search heuristics, yet all share a common limitation that
motivates the regret-based view developed in Section~\ref{sec:regret}.


\subsubsection{Expected Runtime (Hitting Time)}
% - The canonical measure: E[tau_{f*}], expected evaluations to first reach the optimum
% - Equivalent to the expected first hitting time of the target level f*
% - Central to EA theory: Theta(n log n) for (1+1) EA on OneMax,
%   Theta(n^2) for LeadingOnes, Theta(n^k) for Jump_k  [droste2002analysis]
% - Analysis via drift theory, Markov chains, the fitness-level method
%   [doerr2020theory; lengler2019drift]
% - Well-suited for establishing complexity hierarchies and ranking algorithms
%   on a given benchmark
 
The predominant performance measure in the runtime analysis of evolutionary algorithms
is the \textit{expected optimization time}: the expected number of function evaluations
until an optimal solution is queried for the first time~\cite{droste2002analysis,
doerr2020theory}. Formally, for a black-box algorithm producing query sequence
$x_1, x_2, \ldots$, the \textit{first hitting time} of the global optimum is
\[
    \tau_{f^*} \;:=\; \min\bigl\{t \in \mathbb{N} : f(x_t) = f^*\bigr\},
\]
and the expected runtime is $\mathbb{E}[\tau_{f^*}]$. This quantity has proved
immensely useful for establishing complexity hierarchies and ranking algorithms; the
results $\Theta(n \log n)$ for the $(1+1)$~\acrshort{ea} on \textsc{OneMax},
$\Theta(n^2)$ on \textsc{LeadingOnes}, and $\Theta(n^k)$ on $\textsc{Jump}_k$ are
canonical examples~\cite{droste2002analysis}. The primary analytical tools for deriving
such bounds are drift theory~\cite{lengler2019drift}, the fitness-level method, and
Markov chain arguments~\cite{doerr2020theory}.
 
\subsubsection{Fixed-Budget Analysis}
\label{subsubsec:fixed-budget}
% - Orthogonal view: fix T, characterize E[f*_T(A)] = expected quality within budget
%   [jansen2014fixed]
% - Practically motivated: practitioner cannot in general detect optimality;
%   the run terminates at budget exhaustion regardless
% - Different ranking than expected runtime is possible:
%   an algorithm with larger E[tau_{f*}] may still dominate within a given budget
% - Connects directly to simple regret: minimising simple regret <=> maximising E[f*_T]
% - Forward ref: this is the fixed-budget view of performance;
%   §\ref{sec:regret} generalises it to the entire trajectory
 
In \textit{fixed-budget analysis}~\cite{jansen2014fixed}, the evaluation budget $T$ is
fixed in advance and the quantity of interest is the expected quality of the best
solution found within that budget. Denoting by $f^{*}_T(\mathcal{A})$ the best
objective value achieved by algorithm $\mathcal{A}$ within $T$ evaluations, the
fixed-budget framework characterizes $\mathbb{E}[f^{*}_T(\mathcal{A})]$ as a function
of $T$~\cite{jansen2014fixed}. This perspective is practically motivated: in almost all
realistic settings a user cannot detect when the global optimum has been found, and the
algorithm must simply return the best solution found when the budget is
exhausted~\cite{pinto2018practice}. As a consequence, comparing two algorithms under a
fixed budget may reverse the conclusions drawn from comparing their expected runtimes:
an algorithm with a larger $\mathbb{E}[\tau_{f^*}]$ may still produce superior
intermediate solutions within a given budget~\cite{jansen2014fixed}. This framework
connects directly to the simple regret defined in
Section~\ref{subsubsec:sr}: maximizing $\mathbb{E}[f^{*}_T(\mathcal{A})]$ is
equivalent to minimizing expected simple regret at budget $T$.
 
\subsubsection{Fixed-Target Analysis and Runtime Profiles}
\label{subsubsec:fixed-target}
% - Fixed-target: for a quality threshold v, characterise the distribution of tau_v
%   [Doerr & Lengler 2015 - OneMax in black-box models with restrictions]
% - More informative than the single tau_{f*}: a full collection of hitting times
%   across all levels v gives the expectation-based runtime profile  [pinto2018practice]
% - Distributional profile P(v,t) = P(tau_v > t): the primitive object from which
%   E[tau_v] and subsequently cumulative regret are derived
%   (full formal treatment in Chapter X / Runtime Profiles section)
% - Practical value: identifies crossover levels where one algorithm overtakes
%   another, guiding adaptive strategy design  [pinto2018practice]
% - Pinto & Doerr's open question: a combined anytime measure is needed 
 
\textit{Fixed-target analysis} generalizes the expected runtime by considering not only
the global optimum but every intermediate fitness level. For a target quality $v$, the
\textit{first hitting time} is
\[
    \tau_v \;:=\; \min\bigl\{t \in \mathbb{N} : f(x_t) \geq v\bigr\},
\]
and the goal is to characterize the distribution of $\tau_v$ for all relevant $v$.
This viewpoint appears in the analysis of restricted black-box models; for example,
Doerr and Lengler~\cite{doerr2015onemax} derive hitting-time bounds across all
fitness levels for \textsc{OneMax} under several information-theoretic restrictions,
revealing how model constraints shift the difficulty at each tier of the landscape.
 
Pinto and Doerr~\cite{pinto2018practice} formalize this into the \textit{runtime
profile} framework: the \textit{expectation-based runtime profile}
$\bigl(\mathbb{E}[\tau_v]\bigr)_v$ and the strictly finer \textit{distributional
runtime profile} $\mathcal{P}(v,t) := \mathbb{P}(\tau_v > t)$ together capture
the full anytime behavior of an algorithm rather than only its behavior at the endpoint.
Crucially, runtime profiles can identify \textit{crossover fitness levels} where one
algorithm overtakes another in the quality trajectory — behavior invisible to the single
expected optimization time. The formal development, including the connection between
distributional runtime profiles and cumulative regret, is given in
Section~\ref{subsec:regret-profiles}.
 
\subsubsection{Limitations and the Case for a Continuous Performance Measure}
\label{subsubsec:perf-limitations}
% - Expected runtime is binary: found / not found at a single threshold
% - Fixed-budget gives one scalar per budget T, ignores the trajectory
% - Even runtime profiles are a collection of per-level scalars, not a single
%   summary of the entire quality trajectory
% - None of these penalise "wasted" queries on suboptimal regions:
%   an algorithm that stagnates at low fitness and then jumps to the optimum
%   is indistinguishable from one that makes steady progress, by E[tau_{f*}] alone
% - Pinto & Doerr explicitly call for a combined anytime measure from ML/online
%   learning  [pinto2018practice, p.22]; regret is the natural candidate
% - Transition: §2.5 (Sequential Decision-Making) and §2.6 (Regret Formalism)
%   develop this measure from first principles; §2.7 brings it back to BBO
 
Despite their individual strengths, all three frameworks above share a structural
limitation: they measure algorithm performance at isolated thresholds rather than over
the entire quality trajectory. The expected runtime reduces the run to a single hitting
event; fixed-budget analysis reduces it to a single scalar per budget value; and even
the runtime profile, while richer, is a collection of per-level summaries rather than
a unified measure of trajectory quality. Crucially, none of these measures penalizes an
algorithm for spending many evaluations at low fitness values: a strategy that stagnates
near the starting fitness and then discovers the optimum in a single lucky step is
indistinguishable from one that makes steady, monotone progress, as far as
$\mathbb{E}[\tau_{f^*}]$ is concerned.
 
Pinto and Doerr~\cite{pinto2018practice} identify this gap explicitly, calling for
\textit{combined measures} capable of describing the anytime behavior of an evolutionary
algorithm within a unified framework, and naming regret-based measures from machine
learning as a natural candidate, while noting that a fully satisfying formulation
remains an open problem. This project takes that call as its starting point. The
sequential decision-making framework and the formal vocabulary of regret are developed
in Sections~\ref{sec:online}--\ref{sec:regret}; their synthesis with \acrshort{bbo} performance
analysis is the subject of Section~\ref{sec:regret-bbo}.

\newpage
\section{Sequential and Online Optimization Frameworks}
\label{sec:online}

\subsection{Exploration-Exploitation Trade-off}
%   Gathering information (exploration) vs. using current best
%   knowledge (exploitation)
% - Why pure exploitation fails (local optima / suboptimal arms)
% - Why pure exploration fails (never converges to good solutions)
% - This tension is shared with BBO
 
Sequential black-box optimization is governed by the exploration--exploitation trade-off. At each round, the algorithm must either explore uncertain regions of the search space to gain information or exploit the currently best-known region to obtain good values immediately~\cite{cesa2006prediction,bubeck2012regret,lattimore2020bandit}
Since each evaluation is both a measurement and a search step, neither extreme is satisfactory: pure exploitation can lock onto a suboptimal region, while pure exploration wastes budget.

This tension is central to regret-based analysis. Regret measures the cost of not knowing the objective in advance, so it captures both slow identification of good regions and repeated evaluation of poor ones.

\subsection{Multi-Armed Bandits}
% - Formal stochastic bandit model: K arms, reward distributions, T rounds
% - Objective: maximize cumulative reward (equivalently, minimize regret)
% - Stochastic MAB: UCB1, Thompson Sampling — brief intuition  [!CITE!]
% - Contextual bandits: features available at each round  [!CITE!]
% - Pure exploration / best-arm identification: goal is simple regret
%   at the end of the horizon, not cumulative regret  [bubeck2011pure]
% - Connection to BBO: each candidate solution is an "arm"; the oracle
%   is the reward; the algorithm decides which arm to pull
The \acrshort{mab} model is the basic framework for sequential decision-making under partial feedback~\cite{bubeck2012regret,lattimore2020bandit}. In the stochastic $K$-armed setting, an algorithm chooses an arm $A_t \in \{1,\dots,K\}$ at each round and observes only the reward of that arm. If arm $i$ has mean reward $\mu_i$ and $\mu^*=\max_i \mu_i$, the standard pseudo-regret is
\[
    \bar{R}_T = T\mu^* - \sum_{t=1}^{T} \mu_{A_t}
    = \sum_{i:\Delta_i>0} \Delta_i \, \mathbb{E}[N_i(T)],
\]
where $\Delta_i=\mu^*-\mu_i$ is the suboptimality gap and $N_i(T)$ is the number of pulls of arm $i$ up to time $T$~\cite{bubeck2012regret,lattimore2020bandit}. Regret is therefore driven by how often the learner samples suboptimal actions.

This model connects naturally to \acrshort{bbo}. In a finite search space, candidate solutions can be viewed as arms and objective evaluations as arm pulls. For large or structured domains, however, this flat view breaks down, motivating more structured models such as Bayesian optimization.

\subsection{Reinforcement Learning (Overview)}
% - Value functions, policies, and the policy optimization perspective
% - Regret in RL: exploration bonus methods, PAC-MDP bounds  [!CITE!]
% - Note the separation from BBO: RL involves state transitions;
%   BBO is stateless between queries (unless memory is used)
% - Keep brief — RL is context, not a primary focus
 
\subsection{Bayesian Optimization}
% - Surrogate-model approach to expensive BBO  [!CITE Mockus / Shahriari!]
% - Gaussian process surrogate + acquisition function (EI, UCB, PI)
% - Regret in BO: cumulative and simple regret bounds for GP-UCB
%   [Srinivas et al. 2010 — key reference]
% - BO as the bridge domain: it lives at the intersection of online learning
%   and BBO, and has the most mature regret theory outside the bandit literature
% - Distinction from the RSH setting: BO uses a learned surrogate;
%   RSHs are memoryless or use no probabilistic model
\acrfull{bo} treats expensive black-box optimization as a sequential learning problem. The unknown objective is modelled by a probabilistic surrogate, typically a Gaussian process, and an acquisition function selects the next query by balancing predicted value against uncertainty~\cite{jones1998efficient,shahriari2015taking,frazier2018tutorial}.

For this chapter, Bayesian optimization matters mainly because it is the clearest bridge between regret theory and black-box optimization. In particular, \textsc{GP-UCB} admits sublinear cumulative-regret guarantees under Gaussian-process or RKHS\footnote{\textit{abbr.}~Reproducing kernel Hilbert space. (Optionally see \cite{ghojogh2021reproducing})} assumptions~\cite{srinivas2009gaussian,chowdhury2017kernelized,scarlett2017lower}. At the same time, these guarantees rely on surrogate assumptions that are much more natural in low-dimensional continuous spaces than in pseudo-Boolean domains such as $\{0,1\}^n$. This makes regret an established concept in black-box optimization, but not yet one that transfers directly to the combinatorial heuristics studied here.

\newpage
\section{Computational Regret}
\label{sec:regret}

\subsection{Definition of Regret}
% Formal mathematical definition.
% Regret vs. loss.
Let $f : \mathcal{X} \to \mathbb{R}$ be an objective function with global optimum value
\\$f^{*} = \max_{x \in \mathcal{X}} f(x)$. An algorithm that queries points
$x_1, x_2, \ldots, x_T$ incurs \textit{regret} by querying suboptimal solutions. Regret
quantifies the gap between the algorithm's performance and that of an oracle always
querying the optimum~\cite{cesa2006prediction}.
!TODO: add more context with reference to MAB \& rewards!

\subsection{Types of Regret}

\subsubsection{Instantaneous Regret}
% Per-round performance gap.
The \textit{instantaneous regret} at round $t$ is
\[
  \rho_t = f^{*} - f(x_t),
\]
capturing the suboptimality of the single query made at step $t$.
A related variant tracks the \textit{incumbent} (best-so-far) solution rather than
the current query:
\[
  \rho^{\mathrm{inc}}_t = f^{*} - f(\hat{x}_t),
  \quad \hat{x}_t = \operatorname{arg\,max}_{s \leq t} f(x_s).
\]
Because $\hat{x}_t$ is non-decreasing in quality, $\rho^{\mathrm{inc}}_t$ is
monotonically non-increasing over $t$, and is equivalent to simple regret evaluated at time $t$(~\ref{subsubsec:sr}).

\subsubsection{Cumulative Regret}
\label{subsubsec:cr}
% Definition, interpretation, and relationship to IR.
Two cumulative regret variants arise naturally and serve different purposes.
 
The \textit{\acrfull{cr}} after $T$ rounds is~\cite{cesa2006prediction}
\[
  R_T = \sum_{t=1}^{T} \rho_t = \sum_{t=1}^{T} \bigl(f^{*} - f(x_t)\bigr).
\]
This is the standard formulation from the online learning literature: it sums the
instantaneous suboptimality of each individual query, regardless of what the algorithm
has previously observed. Minimizing $R_T$ corresponds to performing well
\textit{throughout} the optimization process, penalizing every suboptimal query equally.
 
The \textit{incumbent cumulative regret} instead tracks the best-so-far process
\[
  B(t) \;:=\; \max_{s \leq t} f(x_s),
\]
yielding
\[
  R^{\mathrm{inc}}_T = \sum_{t=1}^{T} \rho^{\mathrm{inc}}_t = \sum_{t=1}^{T}
  \bigl(f^{*} - B(t)\bigr).
\]
Since $B(t)$ is non-decreasing in $t$, the per-step contributions
$\rho^{\mathrm{inc}}_t$ are non-increasing, and $R^{\mathrm{inc}}_T \leq R_T$ for all
$T$. The incumbent formulation is the appropriate notion when the practitioner retains
the best solution found rather than only the most recent query, which is the standard
convention for the randomized search heuristics studied in this project. All regret
analyses in Chapter~\ref{cha:exper} use $R^{\mathrm{inc}}_T$ unless otherwise stated.
 
In continuous time, the incumbent cumulative regret is approximated via a left-hold
(piecewise-constant) integral
\[
  R^{\mathrm{inc}}_T \approx \sum_{t=1}^{T} \rho^{\mathrm{inc}}_{t-1}\,\Delta t_{\,t},
  \quad \Delta t_{\,t} = t - t_{t-1},
\]
which accounts for unequal spacing between evaluations when function evaluation time is
non-uniform. In all experiments reported here, evaluations are unit-cost and time is
measured in discrete query counts, so $\Delta t_t = 1$ throughout and this reduces to
the standard discrete sum.

\subsubsection{Simple Regret}
% Final recommendation quality.
\label{subsubsec:sr}
The \textit{\acrfull{sr}} after $T$ rounds is~\cite{bubeck2011pure}
\[
  r_T = f^{*} - f(\hat{x}_T),
\]
where $\hat{x}_T = \operatorname{arg\,max}_{t \leq T} f(x_t)$ is the best solution
encountered. Simple regret measures the quality of the \textit{final output} and is the
appropriate notion when the goal is to return a good solution rather than to perform well
during the search. Note that $r_T = \rho^{\mathrm{inc}}_T$: simple regret is the incumbent instantaneous regret at the final round.

When comparing algorithms across multiple independent runs, the standard scalar summary
is the \textit{expected simple regret}
\[
  \mathbb{E}[r_T] = \frac{1}{N}\sum_{i=1}^{N} r_T^{(i)},
\]
where $r_T^{(i)}$ denotes the simple regret of the $i$-th run.

\subsubsection{Normalized Regret}
% Scale-free comparison across problem instances.
To enable comparison across problem instances with different objective scales, simple
regret can be normalized by the range of attainable fitness values:
\[
  \tilde{r}_T = \frac{f^{*} - f(\hat{x}_T)}{f^{*} - f_{\mathrm{worst}}} \in [0, 1],
\]
where $f_{\mathrm{worst}}$ is the worst observed or theoretical fitness value. A value of
$0$ indicates an optimal solution was found; a value of $1$ indicates no improvement over
the worst case. When $f^{*} = f_{\mathrm{worst}}$ (a degenerate constant landscape), the
problem is trivially solved and normalized regret is defined as $0$.





\subsection{Regret Bounds}
% - Upper bounds (algorithmic): what a given algorithm guarantees
%     UCB1: O(sqrt(KT log T)) cumulative regret  [!CITE!]
%     GP-UCB: sublinear cumulative regret for continuous domains  [Srinivas et al.]
% - Lower bounds (information-theoretic): what no algorithm can beat
%     Omega(sqrt(KT)) for K-armed stochastic bandits
% - Minimax regret: min_{algorithm} max_{instance} R_T
%     Connection to Yao's principle  (already introduced in §2.2.2)
% - Instance-dependent (gap-based) bounds: tighter bounds when suboptimality
%     gaps Delta_i = f^* - f(arm_i) are large  [!CITE!]
% - Sublinear regret as the benchmark: R_T = o(T) implies average per-round
%     regret -> 0, i.e., the algorithm learns to behave near-optimally
Regret analysis studies how regret grows with the budget $T$. The central benchmark is sublinear regret: if $R_T=o(T)$, then the average regret $R_T/T$ tends to zero, meaning that the algorithm asymptotically performs as well as an oracle with full information~\cite{cesa2006prediction,bubeck2012regret,lattimore2020bandit}.

In finite stochastic bandits, classic upper bounds show that algorithms such as \textsc{UCB1} achieve logarithmic gap-dependent regret and sublinear worst-case regret~\cite{auer2002finite}. Matching lower-bound results show that logarithmic regret is, in general, unavoidable up to constants under standard assumptions~\cite{lai1985adaptive}. In Gaussian-process bandits, both upper and lower bounds depend on information gain and kernel structure rather than only on the number of actions~\cite{srinivas2009gaussian,chowdhury2017kernelized,scarlett2017lower}. These results are important here mainly as a template: regret can be treated as a complexity measure for sequential optimization.
 
% \subsubsection{Upper Bounds}
 
% \subsubsection{Lower Bounds and Minimax Regret}
 
% \subsubsection{Instance-Dependent Bounds}

\subsection{Connections to Runtime Analysis}
% Relationship between regret and optimization time.
% Trade-offs between exploration and convergence.
% Regret as a refined complexity notion.

\newpage
\section{Regret-Aware Black-Box Optimization}
\label{sec:regret-bbo}
 
% PURPOSE: This section is the intellectual payoff of the chapter.  It
% synthesizes Acts 1 and 2 and positions the project within the literature.
% The reader should finish this section understanding:
%   (a) why regret is a natural and informative measure for BBO,
%   (b) what has already been done at this intersection,
%   (c) what remains open, and
%   (d) how this project advances the state of knowledge.
Regret provides a fine-grained complexity measure in the black-box setting: rather than
simply asking whether an algorithm finds the optimum within $T$ queries, regret tracks
how suboptimal the queries are throughout the process. This is particularly relevant
when every evaluation is costly and performing well before the budget is exhausted
matters in practice.
 
\subsection{Regret as a Performance Measure for Randomized Search Heuristics}
\label{subsec:regret-as-perf}
% - Reframe the classical runtime analysis question in regret terms:
%     Runtime analysis asks: E[T_opt] = ?
%     Regret analysis asks:  E[R_T^inc] = ? for a fixed budget T
% - These are complementary views of the same underlying process
% - Fixed-budget analysis (§2.3.5) is essentially simple regret analysis;
%   making this explicit is conceptually useful
% - Incumbent cumulative regret captures the "area above the fitness curve",
%   generalizing simple regret to the entire trajectory
% - Implications for algorithm design:
%     A low-simple-regret algorithm is good at pure optimization
%     A low-cumulative-regret algorithm is good at "anytime" optimization
%     These objectives can conflict (exploitation vs. exploration tension
%     reappears at the algorithmic design level)
For randomized search heuristics, regret complements rather than replaces runtime analysis. Runtime asks when a target is first reached; regret asks how good the algorithm's queried or retained solutions are throughout the run. Fixed-budget analysis already moves in this direction, since expected simple regret is just the expected final suboptimality after a budget of $T$ evaluations~\cite{jansen2014fixed,bubeck2011pure}.

The distinction matters operationally. Low simple regret is appropriate when only the final returned solution matters. Low cumulative regret is appropriate in anytime settings, where useful intermediate performance also matters~\cite{cesa2006prediction,pinto2018practice}. For the heuristics considered in this project, the incumbent-based formulation is the natural one because the algorithm's output at interruption is the best-so-far solution.


\subsection{Related Work: Regret in Bayesian and Surrogate-Based Optimization}
\label{subsec:rw-bo-regret}
% - BO is the domain with the most developed regret theory for BBO
% - GP-UCB [Srinivas et al. 2010]: O(sqrt(T gamma_T)) cumulative regret
%   where gamma_T is the information gain — first sublinear regret result
%   for a continuous BBO algorithm
% - Thompson Sampling for BO [Russo & Van Roy, Chowdhury & Srinavas]: 
%   Bayesian regret bounds
% - Limitations of BO for the combinatorial setting:
%     GP priors ill-suited to {0,1}^n without kernel engineering
%     Acquisition function optimization is itself an NP-hard BBO problem
%     Poor scaling to large n
% - This motivates studying regret for simpler, combinatorial RSHs
 
\subsection{Related Work: Regret-Inspired Analysis of Evolutionary Algorithms}
\label{subsec:rw-ea-regret}
% - Fixed-budget / anytime guarantees in EA theory
%   [Jansen & Zarges; Doerr & Doerr — heavy-tailed mutations!CITE!]
% - (1+lambda) EA and population diversity as implicit exploration strategies
% - Algorithms explicitly designed with the exploration-exploitation trade-off
%   in mind: parameter control, self-adaptation, hybrid methods
% - Note: most EA theory is runtime-centric; regret framing is novel here
% - What "regret-aware" EA design might look like:
%     Reward-shaping analogies
%     Budget-sensitive mutation rates
%     Acceptance criteria derived from regret minimization objectives
 
\subsection{Related Work: Bandit-Inspired Metaheuristics}
\label{subsec:rw-bandit-meta}
% - Algorithm selection / portfolio methods framed as MAB problems
%   [!CITE Fialho, Maturana, Whitley, etc.!]
% - Operator selection in EAs using UCB / EXP3-style rules
%   [adaptive operator selection literature]
% - Hyper-heuristics with bandit-based move selection
% - Common thread: regret minimization used to select operators or algorithms,
%   not to evaluate the optimizer itself — this project does the latter
 
\subsection{Connections to Runtime Analysis via Drift Theory}
\label{subsec:drift-regret}
% - Drift theory  [lengler2019drift] is the primary tool for EA runtime proofs
% - Additive, multiplicative, variable drift theorems
% - How drift on the fitness function implies bounds on E[R_T^inc]:
%     A lower-bounded drift implies the incumbent fitness increases quickly
%     => the cumulative regret decreases quickly
%     Formalizing this connection is a contribution of this project
% - This subsection can be brief if the formal derivation belongs in
%   Chapter 3 (Methodology / Theory); just sketch the intuition here
 
\subsection{Open Questions and Project Scope}
\label{subsec:open-questions}
% - Summary of what is NOT yet known:
%     Tight cumulative regret bounds for (1+1) EA / RLS on standard benchmarks
%     How the exploration-exploitation tension manifests quantitatively
%     Whether epsilon-greedy and SA schedules can be tuned to minimize regret
%       rather than (only) expected runtime
% - Research questions this project addresses:
%     [enumerate your specific RQs here]
% - Chapter roadmap: what follows from this background
