% Word count aim: 1500-2000
\chapter{Introduction}
\label{cha:intro}

% Put in an illustrative picture
\section{Motivation}
Across engineering, the natural sciences, and machine learning, a
recurring class of optimization problems resists conventional
analytical treatment. The objective function may arise from a physical
experiment, a high-fidelity simulation, or a complex computational
pipeline whose internal structure is either inaccessible or too
intricate to exploit.
In such settings, the practitioner's only recourse is to query the
objective at a candidate solution and observe the resulting value;
no gradient, curvature, or closed-form expression is
available~\cite{liu_primer_2020}. Problems of
this kind are collectively termed \textit{\acrfull{bbo}} problems.
Industrial process optimization, drug-design pipelines, and the tuning of
machine-learning models~\cite{bajaj2021black,bischl2017mlrmbo} belong
to this class of problems.

\begin{figure}[H]
  \centering
  \begin{tikzpicture}[
      node distance=2.8cm,
      box/.style={rectangle, rounded corners, draw, align=center,
      minimum width=3.0cm, minimum height=1.0cm},
      arrow/.style={->, thick}
    ]
    \node[box] (alg) {Black-box\\algorithm $\mathcal{A}$};
    \node[box, right=of alg] (oracle) {Function-evaluation\\oracle};
    \node[box, below=of oracle] (obj) {Unknown
    objective\\$f:\{0,1\}^n\to\mathbb{R}$};

    \draw[arrow] (alg) -- node[above] {query $x_t$} (oracle);
    \draw[arrow] (oracle) -- node[below] {value $f(x_t)$} (alg);
    \draw[arrow] (oracle) -- (obj);
  \end{tikzpicture}
  \caption{Black-box optimization interaction model. The algorithm receives
    only zeroth-order objective values and must adapt future queries from the
  observed history.}
  \label{fig:bbo-query-loop}
\end{figure}

Without structural access, practitioners turn to general-purpose
\textit{\acrfull{rsh}}s: stochastic algorithms such as Evolutionary
Algorithms (\acrshort{ea}) and Simulated Annealing (\acrshort{sa})
that search by iterative perturbation and
selection~\cite{abdel2018metaheuristic,rajwar2023exhaustive}. The
standard theoretical question for such algorithms is the number of
evaluations needed to first sample the optimum. This single
metric (the expected runtime) has dominated the field for decades
and underpins a rich body of complexity
results~\cite{droste2002analysis,doerr2020theory,lengler2019drift}.

However, expected runtime is a blunt instrument. It says nothing
about solution quality \textit{before} the optimum is reached.
In practice, budgets are finite, runs are interrupted, and the
best-so-far solution is what actually gets used~\cite{jansen2014fixed}.
An algorithm that stagnates and then gets lucky looks identical, by
this measure, to one that makes steady progress throughout. This gap
has been noted explicitly in the literature: Pinto and
Doerr~\cite{pinto2018practice} call for \textit{combined anytime
measures} from machine learning, naming regret as a natural candidate
while acknowledging that a satisfying formulation remains open.

This project answers that call. By connecting the \textit{regret}
framework from online learning~\cite{cesa2006prediction,lattimore2020bandit}
to the \textit{runtime profile} framework from evolutionary
computation~\cite{pinto2018practice}, a unified anytime measure
is established — one that captures solution quality across the
\textit{entire} run, not only at its end. A bespoke experimental
framework is developed to validate this connection and to investigate,
across a broad range of benchmark problems and algorithm families,
what regret trajectories reveal about landscape structure that
classical measures cannot.

% Many aspects of our day-to-day lives revolves around finding
% optimality in problems that we can have no comprehensible insight
% about. This abstraction could be due to several factors. For
% instance, optimizing a design value in many large-scale and complex
% systems proves computationally or materially expensive. It can be
% observed in industrial processes of !TODO!. These
% types of problems are said to be \acrfull{bbo} problems.

% We do not possess many tools to interact with them save from a
% cost/objective function $\mathbf{f}: \mathcal{X} \mapsto \mathcal{S}$
% (the `oracle') that provides an abstract representation of the
% problem, where $\mathcal{X}$ is the search space, $\mathcal{S}$ the
% evaluation space !TODO: find concrete refs!. Typically, $\mathcal{S}
% \subseteq R$. The task is to minimize or maximize the solutions
% produced by this objective function, i.e., find a solution that Much
% study is done in optimizing this cost function without the liberty of
% classic mathematical tools such as continuity, smoothness and
% convexity (these methods warrant the oracle to return additional
%   information, such as gradients, apart from the evaluation value of an
% input \cite{liu_primer_2020}).

% While this approach secures the bounds on the first sampling of the
% optimum, it presents no clues on the solution quality produced by the
% algorithm. Consider the case when the algorithm samples the optimum
% at time $t$, then in the next step $t+1$, moves completely away from
% the vicinity. This algorithm might then produce worse answers, asymptotically.

% \subsection{Conventional Solutions}
% % A brief history of how this problem came to be
% Classical analysis of solving \acrshort{bbo} problems are concerned
% with the so-called \acrfull{bbc}, that is, deriving the runtime
% bounds for an algorithm to sample a (global) optimum, depending only
% on the problem size $\mathbf{n}$. The reasoning behind this is
% palpable; the \textit{computational} complexity of evaluating the
% objective function for such \acrshort{bbo} systems are usually expensive.

% A trivial example is Hyperparameter Tuning in \acrfull{ml}
% models~\cite{bischl2017mlrmbo}. Searching for optimal hyperparameters
% is a \acrfull{bbo} problem that
% optimizes a single or multiple objectives on the provided (training
% and validation) data.
% (e.g., accuracy, F1-score, $R^2$ score.).

% Give context to the reader about the objectives

\begin{longtable}{||p{0.48\textwidth}|p{0.48\textwidth}||}
  \hline
  \textbf{Broader Aims} & \textbf{Specific Objectives} \\
  \hline
  \hline
  \endhead
  \hline
  \textbf{Aim 1. } \emph{Consolidate the theoretical foundations of
    \acrshort{bbo}, including the standard performance measures and
  their limitations.}
  &
  \begin{enumerate}[topsep=0pt, partopsep=0pt, itemsep=0pt,
      parsep=0pt, leftmargin=*]
    \item Formalize the expected runtime, fixed-budget, and
      fixed-target frameworks.
    \item Identify the shared limitation: the absence of a combined
      anytime performance measure.
    \item Review the regret vocabulary from online learning and
      situate it relative to \acrshort{bbo} complexity theory.
  \end{enumerate}
  \\
  \hline
  \textbf{Aim 2. } \emph{Explore a viable relationship between the
  canonical measure of \acrshort{bbo} algorithms and regret.}
  &
  \begin{enumerate}[topsep=0pt, partopsep=0pt, itemsep=0pt,
      parsep=0pt, leftmargin=*]
    \item Define the distributional runtime profile
      $\mathcal{P}(v, t) = \mathbb{P}(\tau_v > t)$.
    \item Prove the tail-sum regret representation identity
      (Theorem~\ref{thm:regret-profile}).
    \item Derive the asymptotic corollary connecting long-run regret
      to the expected hitting times $\mathbb{E}[\tau_v]$.
  \end{enumerate}
  \\
  \hline
  \textbf{Aim 3. } \emph{Develop a reproducible experimental framework
  suited to trajectory-level regret analysis.}
  &
  \begin{enumerate}[topsep=0pt, partopsep=0pt, itemsep=0pt,
      parsep=0pt, leftmargin=*]
    \item Implement the \texttt{regret} framework with declarative
      \acrshort{yaml}-driven experiment dispatch.
    \item Support full trajectory recording, regret metric computation,
      and runtime profile analysis.
    \item Validate the implementation via a test-driven development
      discipline with enforced coverage distribution.
  \end{enumerate}
  \\
  \hline
  \textbf{Aim 4. } \emph{Investigate empirically how regret trajectories
  reflect landscape structure across a broad benchmark suite.}
  &
  \begin{enumerate}[topsep=0pt, partopsep=0pt, itemsep=0pt,
      parsep=0pt, leftmargin=*]
    \item Verify the tail-sum identity numerically on tractable benchmarks.
    \item Characterize the diagnostic signatures of barrier-crossing,
      epistatic ruggedness, and hierarchical difficulty in cumulative
      regret trajectories.
    \item Identify conditions under which classical measures (\acrshort{ttfo},
      $P(\text{opt})$) are uninformative and regret trajectories remain
      structured.
  \end{enumerate}
  \\
  \hline
  \caption{Project aims and specific objectives.}
  \label{tab:aims}
\end{longtable}

\subsection{Success Criteria}
\label{subsec:intro:success}
Following this project's intended aims (Table~\ref{tab:aims}), the
\emph{success criteria} for these achievements are quantified in
Table~\ref{tab:succrit}.

\begin{longtable}{@{}p{0.15\linewidth}p{0.79\linewidth}@{}}

  \toprule
  \textbf{Area} & \textbf{Criteria} \\
  \midrule
  \endfirsthead

  \toprule
  \textbf{Area} & \textbf{Criteria} \\
  \midrule
  \endhead

  % \bottomrule
  % \endlastfoot

  \textsc{Theory} &
  The tail-sum identity (Theorem~\ref{thm:regret-profile}) must be both
  proved rigorously \textit{and} corroborated numerically to within
  Monte-Carlo sampling error on at least two benchmark functions. Proof
  alone is insufficient: if the identity cannot be verified in the
  finite, discrete experimental setting, it provides no practical basis
  for the empirical analyses that follow. Agreement between
  trajectory-derived and profile-derived expected \acrfull{cr}
  curves procured from experiments constitutes the verification. \\
  \hline

  \textsc{Empirics} &
  \acrfull{cr} trajectories must yield qualitatively distinct
  signatures across at least three landscape archetypes (for instance,
  unimodal, barrier, and deceptive) that are not separated by
  expected runtime (time to optimum) alone. This is the
  operational test of the project's central claim that regret provides
  strictly more diagnostic information than classical measures. Without
  this, the theoretical contribution, though formally correct, would
  have no demonstrated practical utility.\\
  \hline

  \textsc{Software} &
  The \texttt{regret} framework must execute all 13 experimental suites
  reproducibly and maintain a minimum of 50\% branch coverage enforced
  in \acrshort{ci}. This threshold is the evidentiary floor for the
  empirical contribution: results obtained from an unvalidated
  implementation cannot support claims about algorithm behavior, and
  reproducibility is a precondition for any comparative conclusion.\\
  \hline

  \caption{Success criteria for the project}
  \label{tab:succrit}

\end{longtable}

\section{Contributions}
\label{sec:intro:contributions}

This project makes the following principal contributions, as
summarized in Table~\ref{tab:contributions}.
\begin{longtable}{@{}p{0.15\linewidth}p{0.79\linewidth}@{}}

  \toprule
  \textbf{Area} & \textbf{Contributions} \\
  \midrule
  \endfirsthead

  \toprule
  \textbf{Area} & \textbf{Contributions} \\
  \midrule
  \endhead

  % \bottomrule
  % \endlastfoot

  \textsc{Theory} &
  Established a formal connection between expected incumbent
  \acrfull{cr} and
  the distributional runtime profile of~\citet{pinto2018practice}
  (Theorem~\ref{thm:regret-profile}), yielding the \textit{combined anytime
  measure} identified as an open problem in the evolutionary computation
  literature. Corollary~\ref{cor:asymptotic-regret} relates infinite-horizon
  regret to the sum of expected first hitting times over all fitness levels,
  bridging expected-runtime analysis and regret-based evaluation. The
  results hold
  for any randomized black-box algorithm on any finite pseudo-Boolean domain. \\
  \hline

  \textsc{Empirics} &
  Introduced a structured suite of 13 experiments that numerically validates the
  tail-sum identity (Suite~08) and studies how regret trajectories reflect
  landscape structure, including barrier crossings (Suites~04--06, 09--10),
  epistatic ruggedness (Suites~03, 12), hierarchical difficulty (Suite~07), and
  scaling behavior (Suite~13). The results show that \acrshort{cr} captures
  diagnostic information about trapped phases and saturation that
  fixed-budget and
  expected-runtime measures miss when $P(\mathrm{opt}) \approx 0$. \\
  \hline

  \textsc{Software} &
  Developed \texttt{regret}, a declarative, reproducible, and parallelized
  experimental framework for pseudo-Boolean and combinatorial optimization
  problems, with support for trajectory recording, regret metrics,
  runtime-profile
  analysis, and structured output in \acrshort{json} and
  \acrshort{pdf} formats. \\
  \hline

  \caption{Main contributions from the project.}
  \label{tab:contributions}
\end{longtable}

\newpage
\section{Report Structure}
This report is structured as follows:
\paragraph{Background [Chapter~\ref{cha:bg}]} compiles the necessary
knowledge sufficient to grasp the intent and work presented in the later texts.
\paragraph{Runtime Profiles and Anytime Performance
[Chapter~\ref{cha:profiles}]} delves into the core theory developed
this project, coupling a measure of regret with the metric of
`Runtime Profiles'.
\paragraph{Supplementary Software Design [Chapter~\ref{cha:design}]}
documents the novel, supplementary experimental frameworks developed
for this project. This software facilitated the experimentation and
exploratory analysis of various standard \acrshort{bbo} problems
\paragraph{Methodology [Chapter~\ref{cha:exper}]} inscribes the
exploratory and verification (based on concepts in
Chapter~\ref{cha:profiles}) experiments coordinated in this project;
emphasis is provided on specific rationale and decisions taken for
motivating the exploration.
\paragraph{Evaluation and Discussion of Results
[Chapter~\ref{cha:results}]} confers the results gathered from the
experiments conducted. Pointwise regret metrics extracted from the
experiments (against known optimums) are analyzed qualitatively and
quantitatively (especially for the verification experiments mentioned earlier).
\paragraph{Conclusion and Future Work [Chapter~\ref{cha:conc}]}
collates the substance, contributions, and achievements of the
project. Reflections and potential future directions beyond the
seminal work in this project are mapped out.

% \subsection{Open Questions and Project Scope}
% \label{subsec:open-questions}
% % - Summary of what is NOT yet known:
% %     Tight cumulative regret bounds for (1+1) EA / RLS on standard benchmarks
% %     How the exploration-exploitation tension manifests quantitatively
% %     Whether epsilon-greedy and SA schedules can be tuned to minimize regret
% %       rather than (only) expected runtime
% % - Research questions this project addresses:
% %     [enumerate your specific RQs here]
% % - Chapter roadmap: what follows from this background

% Despite the maturity of both runtime analysis and regret theory as individual
% disciplines, their intersection remains largely unexplored. The
% principal open questions
% motivating this project are as follows.

% \paragraph{Tight cumulative regret bounds for standard \acrshort{rsh}s.}
% Closed-form expressions for $\mathbb{E}[\mathrm{CR}(T)]$ are not yet
% established for
% the $(1+1)$-\acrshort{ea} or \acrshort{rls} on the canonical
% pseudo-Boolean benchmarks.
% While the tail-sum identity of Theorem~\ref{thm:regret-profile}
% provides a route from
% known hitting-time bounds to regret bounds, the resulting expressions
% are in general
% loose. Obtaining tight asymptotic characterizations of
% $\mathbb{E}[\mathrm{CR}(T)]$
% on \textsc{OneMax}, \textsc{LeadingOnes}, and \textsc{Jump} is an
% open theoretical
% problem.

% \paragraph{Quantitative exploration--exploitation trade-offs.}
% The $\varepsilon$-greedy and decaying-$\varepsilon$-greedy
% \acrshort{rls} variants
% (Section~\ref{subsubsec:ea}) introduce explicit exploration rates,
% but the optimal
% schedule for minimizing cumulative regret rather than expected
% runtime is unknown.
% It is unclear whether the decay schedule $\varepsilon_t =
% \varepsilon_0 / t$ that
% mirrors classical bandit theory~\cite{auer2002finite} is appropriate
% in the pseudo-Boolean
% setting, where the reward structure is not stationary.

% \paragraph{Regret as a landscape diagnostic.}
% Cumulative regret trajectories have a characteristic shape that reflects the
% structure of the fitness landscape: a linear ramp during a trapped
% phase, a sharp
% drop upon barrier crossing, and a flat floor after convergence. Whether these
% signatures are sufficiently distinct across barrier types (Jump,
% Trap, Plateau) to
% serve as practical diagnostics, and whether they remain informative when the
% optimum is never reached within the budget, has not been studied
% systematically.

% This project addresses these questions empirically. The experimental suite of
% Chapter~\ref{cha:exper} provides trajectory-level evidence for the
% landscape diagnostic
% value of cumulative regret across all three barrier archetypes, and
% the runtime profile
% verification of Suite 08 confirms the tail-sum identity of
% Theorem~\ref{thm:regret-profile} numerically. Tight analytical bounds and the
% derivation of regret-optimal parameter schedules are identified as
% directions for
% future work in Section~\ref{cha:conc}.
%
% \newpage
% \section{Project Outline}
% TODO: Maybe include Gantt Chart...?
% Sem 1
%
% \subsection{Literature Study}

% \subsection{Isolating Relevant Benchmarks}

% \subsection{Development of Experiment framework}

% \subsection{Orchestrating Experiments}

% \subsection{Evaluating Results}
