% \chapter{Experiments}
% \label{cha:exper}
%
% \section{Experiment Structure}
%
% \section{Exploratory Analysis of Regret \& Runtime}
% \subsection{Experiment 1}
% \subsection{Experiment 2}
% \subsection{Experiment 3}
% \subsection{Experiment 4}
%
%
% \section{Expected Cumulative Regret and Runtime Profiles}
% % TODO: Clarify structure of such experiments
% % Gist: Math motiv -> setup -> collection of results (mentioned for
% both the experiment types in next chapter)
% \subsection{}

\chapter{Experiments}
\label{cha:exper}

\section{Experiment Structure}
\label{sec:exper:structure}

All experiments share a common infrastructure defined in the YAML configuration
files under \texttt{configs/experiments/} !TODO: Cite appendices!.
Each configuration specifies four top-level blocks: \texttt{suite}
(global settings), \texttt{problems} (benchmark functions),
\texttt{algorithms} (optimizers), and \texttt{plotting} (output).

\subsection{Repetitions and budget}
Each algorithm--problem pair is run for $R$ independent repetitions
(typically 25--100) at budgets drawn from a per-suite budget list. The
display horizon \texttt{budget\_for\_plots} is chosen to be roughly twice the
expected time-to-first-optimum of the hardest algorithm on each problem, so
trajectory plots show both approach and post-convergence behavior.

\subsection{Execution mode}
All suites reported in this chapter were run in \texttt{full} mode with
parallelism enabled. Full mode records complete per-evaluation trajectories,
which are required for the instantaneous-regret, cumulative-regret, and
\acrshort{ttfo} analyses, and for all runtime-profile computations.

\subsection{Visualization and Table Strategy}
\label{subsec:exper:vis}

Each suite produces plots and summary tables from the stored JSON results.
Table~\ref{tab:plot-types-cat} lists every plot type with its key, diagnostic
purpose, and the suites or conditions that enable it.

\small
\begin{longtable}{@{}llp{4.6cm}l@{}}
  \toprule
  \textbf{Group} & \textbf{Output key} & \textbf{Diagnostic purpose}
  & \textbf{$y$-axis} \\
  \midrule
  \multirow{3}{*}{Aggregate}
  & \texttt{regret\_curves}          & Mean simple regret
  vs.\ budget; crossover budget between algorithms & log \\
  & \texttt{convergence\_probability}& $P(\text{opt})$ vs.\ budget;
  separates partial from full convergence & linear \\
  & \texttt{comparison\_heatmap}     & $\log_{10}(\text{mean
  regret})$ colour-coded over all alg--budget cells & --- \\
  \midrule
  \multirow{2}{*}{Distribution}
  & \texttt{regret\_boxplots}        & Run-level spread of simple
  regret; skewness from non-converging runs & linear \\
  & \texttt{performance\_profile}    & Empirical CDF of regret on log
  axis; steep rise = reliable convergence & log \\
  \midrule
  \multirow{5}{*}{Trajectory}
  & \texttt{history\_best}           & Monotone incumbent fitness
  with $f^{*}$ reference & linear \\
  & \texttt{regret\_instantaneous}   & IR of evaluated solution;
  fluctuates when worsening moves accepted & log \\
  & \texttt{regret\_instantaneous\_best} & IR of incumbent (monotone
  non-increasing); non-zero floor = saturation & log \\
  & \texttt{regret\_cumulative}      & Integral of current-solution
  IR; penalises slow convergence & linear \\
  & \texttt{regret\_cumulative\_best}& Integral of incumbent IR; key
  signal in barrier-crossing suites & linear \\
  & \texttt{ttfo\_distribution}      & Jittered TTFO samples per
  algorithm; conditioned on convergence & linear \\
  \midrule
  \multirow{3}{*}{\parbox{1.6cm}{Runtime\\profile\\(suites\\03, 08, 12)}}
  & \texttt{inverse\_runtime\_profile\_surface} & $P(\tau_v \leq T)$
  heat map over $(T,v)$; saturation frontier visible & --- \\
  & \texttt{inverse\_runtime\_profile\_curves}  & $P(\tau_v \leq T)$
  vs.\ $T$ at $f^{*}\times\{0.25,0.5,0.75,0.95\}$ & linear \\
  & \texttt{cr\_profile\_verification}          & Overlay of direct
  and profile-derived $\mathbb{E}[\mathrm{CR}(T)]$ & linear \\
  \midrule
  Summary & per-problem table            & Mean, median, SD, IQR,
  95\% CI, $P(\text{opt})$, run count per algorithm & --- \\
  & aggregate table              & Across-problem mean regret, mean
  $P(\text{opt})$, best-count per algorithm & --- \\
  \bottomrule
  \caption{Plot types produced by the experimental pipeline. The ``Output key''
    column matches the YAML \texttt{plots} sub-block name.
    All trajectory and distribution plots are generated at
    \texttt{budget\_for\_plots}; aggregate plots pool across all budgets in the
  suite list. IR = instantaneous regret; CR = cumulative regret.}
  \label{tab:plot-types-cat}
\end{longtable}

For hill-climbing algorithms (RLS, $(1+1)$-EA) the current-solution and
incumbent IR variants coincide because no worsening moves are accepted; for SA
and \acrshort{rls}Exploration they diverge during phases of active exploration,
making the current-solution variant a direct diagnostic of temperature-driven
behaviour. Cumulative regret is always plotted on a linear $y$-axis: its slope
during the trapped phase grows linearly with time, and a log axis would suppress
that signal.

\subsection{Suite Numbering and Roles}

Suites are numbered 01--14. Table~\ref{tab:suite-overview} gives a complete
reference; the methodological context for runtime-profile suites (03, 08, 12)
is developed in Chapter~\ref{cha:profiles}.

\begin{table}[hp]
  \centering
  \small
  \setlength{\tabcolsep}{4pt}
  \rotatebox{90}{
    \begin{tabular}{@{}cp{4cm}p{3.5cm}lrrl@{}}
      \toprule
      \textbf{Suite} & \textbf{Role} & \textbf{Problems} & \textbf{Algorithms}
      & \textbf{$R$} & \textbf{Max budget} & \textbf{Profile} \\
      \midrule
      01 & Baseline comparison
      & 6 pseudo-boolean
      & RLS, RLSExpl, EA, SA-Log-d2.0
      & 50 & 10\,000 & No \\
      02 & Extended benchmark
      & 9 pseudo-boolean + combinatorial
      & same 4 families (with overrides)
      & 50 & 20\,000 & No \\
      \midrule
      08 & Tail-sum verification
      & OneMax, LeadingOnes
      & RLS, $(1+1)$-EA
      & 100 & 10\,000 & \textbf{Yes} \\
      \midrule
      03 & NK ruggedness ($6\times5$ grid)
      & 5 NK ($k\in\{1,2,4,6,8\}$, $n=20$)
      & RLS, EA, 6$\times$SA-Log
      & 25 & 50\,000 & \textbf{Yes} \\
      04 & Trap: deception width $\times$ rate
      & 4 Trap ($k\in\{5,8,10,20\}$)
      & 4$\times$EA + SA-Log-d4.0
      & 25 & 10\,000 & No \\
      05 & Jump: mutation rate / plateau onset
      & 5 Jump ($k\in\{2,3,5,7,10\}$)
      & 3$\times$EA (fixed + adaptive $k/n$)
      & 25 & 20\,000 & No \\
      06 & Plateau: neutral walk $\times$ rate
      & 4 Plateau ($k\in\{2,5,10,20\}$)
      & 3$\times$EA + SA-Log-d4.0
      & 30 & 20\,000 & No \\
      07 & \acrshort{hiff} scaling
      & 3 \acrshort{hiff} ($n\in\{32,64,128\}$)
      & RLS, 4$\times$EA, SA-Log-d4.0
      & 25 & 50\,000 & No \\
      09 & Jump: plateau duration (extended)
      & 5 Jump (same as 05)
      & 2$\times$EA (fixed + adaptive)
      & 25 & 50\,000 & No \\
      10 & Barrier archetype comparison
      & Jump/Trap/Plateau ($k=5$, $n=100$)
      & 4$\times$EA ($0.5/n$--$5/n$)
      & 30 & 50\,000 & No \\
      11 & SA cooling diagnostics
      & OneMax, LeadingOnes
      & RLS, 6$\times$SA-Log, SA-Lin, SA-Exp
      & 50 & 10\,000 & No \\
      12 & NK saturation
      & 5 NK ($k\in\{0,2,4,6,8\}$, $n=20$)
      & RLS, $(1+1)$-EA, SA-Log-d4.0
      & 25 & 50\,000 & \textbf{Yes} \\
      13 & Regret scaling with $n$
      & 4 OneMax ($n\in\{25,50,100,200\}$)
      & $(1+1)$-EA at rate $1/n$
      & 60 & 10\,000 & No \\
      14 & Population dynamics
      & OneMax, LeadingOnes, TwoMax, Jump-3
      & $(1+1)$-EA + 4$\times(\mu{+}\lambda)$-EA
      & 30 & 10\,000 & No \\
      \bottomrule
    \end{tabular}
  }
  \caption{Overview of all 14 experimental suites.
    $R$ = runs per algorithm--problem--budget combination.
    ``Profile'' indicates whether \texttt{suite.profile:~true} is set.
    Budget values are in evaluations; ``max'' is the largest value in the
  suite budget list.}
  \label{tab:suite-overview}
\end{table}

\section{Baseline Suites}
\subsection{Suite 01: Algorithm-Family Comparison}

Suite 01 establishes a reference comparison of four algorithm families across
six pseudo-boolean problems of increasing structural difficulty: \acrfull{rls},
\acrshort{rls}Exploration, the $(1+1)$-\acrshort{ea}, and \acrshort{sa} with
logarithmic cooling at $d=2.0$. SA is reduced to a single representative
schedule here; the effect of varying the schedule parameter is deferred to
suites 03 and 11.

The six problems are: OneMax ($n=100$), LeadingOnes ($n=100$), TwoMax
($n=100$), Jump-3 ($n=100$, $k=3$), Trap-10 ($n=100$, $k=10$), and BinVal
($n=50$). BinVal uses $n=50$ rather than $n=100$ because its exponential
fitness structure already produces large fitness gaps and rapid convergence at
this size; the \texttt{budget\_for\_plots} of 500 reflects this. The 50 runs
per combination are evaluated at six budgets: $\{200, 500, 1000, 2000, 5000,
10000\}$. Pairwise statistical annotations use \acrshort{rls} as the reference
algorithm on box-plots, performance profiles, and \acrshort{ttfo} distributions.

\subsection{Suite 02: Extended Pseudo-Boolean Benchmark}

Suite 02 extends the baseline to nine problems, adding Jump-5, Trap-5,
Plateau-10, \acrshort{hiff}-64, and a combinatorial instance (Petersen-3-Color).
The algorithm set is identical to suite 01. Budgets range from 500 to 20\,000
over 50 runs. Profile analysis is disabled.

Three problems receive problem-specific algorithm overrides (full parameter
  values are listed in Table~\ref{tab:suite02-overrides} in the
appendix). For \acrshort{rls}Exploration on Jump-3, Jump-5, and Trap-5, the
exploration rate is raised to $\varepsilon=0.05$ with decay disabled,
maintaining
persistent non-improving moves on barrier problems. For SA, the cooling
parameter is tripled to $d=6.0$ on Jump-5 and Trap-5, raising acceptance
probability for worsening moves throughout the run. Both modifications target
the same objective: improving escape from local or deceptive optima at moderate
budget.

\begin{table}[H]
  \centering
  \caption{Problem-specific algorithm overrides in suite 02. For
    \acrshort{rls}Exploration the default is $\varepsilon=1/n=0.01$
    with exponential
    decay ($\varepsilon_t = \varepsilon/t$); for SA the default cooling
  parameter is $d=2.0$.}
  \label{tab:suite02-overrides}
  \small
  \begin{tabular}{@{}lllll@{}}
    \toprule
    \textbf{Algorithm} & \textbf{Problem} & \textbf{Parameter} &
    \textbf{Default} & \textbf{Override} \\
    \midrule
    \multirow{3}{*}{\acrshort{rls}Exploration}
    & Jump-3  & $\varepsilon$, decay & $0.01$, on & $0.05$, off \\
    & Jump-5  & $\varepsilon$, decay & $0.01$, on & $0.05$, off \\
    & Trap-5  & $\varepsilon$, decay & $0.01$, on & $0.05$, off \\
    \midrule
    \multirow{2}{*}{SA-Log-d\_2.0}
    & Jump-5  & $d$ & $2.0$ & $6.0$ \\
    & Trap-5  & $d$ & $2.0$ & $6.0$ \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Verification Suite}

\subsection{Suite 08: Runtime Profile Verification}

Suite 08 validates the tail-sum identity
\[
  \mathbb{E}[\text{CR}(T)]
  = \sum_{v=1}^{f^{*}} \sum_{t'=1}^{T}
  \bigl[1 - P(\tau_v \leq t')\bigr]
\]
on OneMax and LeadingOnes ($n=100$) using \acrshort{rls} and the $(1+1)$-EA.
With $R=100$ independent runs (the highest run count in the study) Monte-Carlo
variance in $\widehat{\mathbb{E}}[\text{CR}(T)]$ is suppressed to the point
where any visible gap between trajectory-derived and profile-derived curves in
the verification plot can be attributed to numerical discretization rather than
sampling noise. The \texttt{budget\_for\_plots} for LeadingOnes is set to the
maximum available (10\,000) to capture the full cumulative regret ramp, which
extends further than on OneMax due to its $\Theta(n^{2})$ expected convergence
time. Profile analysis is enabled; pairwise annotations use \acrshort{rls} as
the reference.

\section{Exploratory Analysis of Regret and Runtime}

The exploratory suites address four themes: barrier structure and mutation rate
(suites 04--06, 09--10), NK ruggedness and SA cooling (suites 03, 11--12),
hierarchical structure and scaling (suites 07, 13), and population dynamics
(suite 14).

\subsection{Suite 03: NK Landscape Ruggedness}

% Suite 03 crosses six SA cooling schedules with five NK landscape instances at
% different epistasis levels ($k \in \{1, 2, 4, 6, 8\}$, $n=20$, seed 42).
% \acrshort{rls} and the $(1+1)$-\acrshort{ea} are included as
% zero-temperature baselines;
% without a baseline that performs no temperature-driven acceptance, it would
% be impossible to determine whether any SA variant improves on pure
% hill-climbing. On low-epistasis instances ($k=1$, $k=2$) SA may in fact
% underperform hill-climbing due to temperature-induced noise, which the
% baselines make visible.

% The six SA variants use logarithmic cooling with $d \in
% \{0.5, 1.0, 2.0, 4.0, 8.0, 16.0\}$. Critically, all SA entries carry
% empty \texttt{by\_problem} blocks. This is a deliberate choice: the same $d$
% value applies across all five NK instances, so the suite is a full
% $6 \times 5$ grid of (cooling parameter, epistasis level) pairs with no
% per-cell tuning. Any interaction visible in the results therefore reflects
% the structure of the problem rather than instance-specific calibration.

% The $n=20$ constraint is imposed by the NKLandscape implementation: computing
% the exact optimum requires exhaustive search over $2^{n}$ bitstrings, which
% is only feasible at this size. The problem seed is fixed at 42 across all five
% instances, so the randomness in each landscape is determined solely by $k$;
% the same seed is reused in suite 12, making the two NK suites directly
% comparable on the same landscape realisations.

% Profile analysis is enabled, producing inverse runtime profile surfaces (one
% per algorithm), profile curves at selected fitness levels, and the profile
% verification plot. The extended budget list, reaching 50\,000, is required
% because high-epistasis NK instances with conservative cooling schedules (large
% $d$) may not reach their saturation floor until late in the run.

Suite 03 crosses six logarithmic SA schedules ($d \in
\{0.5,1.0,2.0,4.0,8.0,16.0\}$) with five NK instances ($k \in \{1,2,4,6,8\}$,
$n=20$, seed 42). \acrshort{rls} and the $(1+1)$-\acrshort{ea} are included as
zero-temperature baselines so that the effect of temperature-driven acceptance
can be separated from plain hill-climbing; on low-epistasis instances SA may in
fact underperform hill-climbing, which the baselines make visible.

All SA entries carry empty \texttt{by\_problem} blocks, forming a clean
$6 \times 5$ grid over cooling parameter and epistasis level without per-problem
tuning. The restriction to $n=20$ is imposed by the exact-optimum computation
used in the NK implementation. Profile analysis is enabled, and the budget
extends to 50\,000 because high-epistasis instances may approach saturation only
late in the run.

\subsection{Suite 04: Trap Function Mutation Rate}

% Suite 04 places the Trap function under four $(1+1)$-\acrshort{ea}
% mutation rates and one
% SA variant across four deception widths ($k \in \{5, 8, 10, 20\}$, $n=100$).
% The four \acrshort{ea} entries use rates $0.5/n$, $1/n$, $2/n$, and
% $5/n$ with no
% per-problem overrides; the same rate applies to all four Trap instances.

% The SA entry (SA-Log-d\_4.0) uses $d=4.0$ as the default and overrides to
% $d=8.0$ on Trap-k20 only. This is the only problem-specific override in the
% suite, applied to the widest deception width. The override doubles the
% temperature at every evaluation step, implying that at $k=20$ the default
% schedule was judged insufficient to maintain adequate acceptance probability
% across the wider deceptive basin.

% Trap-k1 (fully deceptive with $k=1$) is excluded. At $k=1$ the trap has no
% hill-climbing gradient leading toward the optimum at any scale. The fitness
% function decreases monotonically from the all-zeros attractor to the
% all-ones optimum for all $u < n$. Hence, all algorithms converge to
% zero on all
% metrics. This produces uninformative flat plots that consume run
% budget without
% generating insight. The gap from $k=5$ to $k=8$ (skipping $k=3$ and $k=6$,
% $k=7$) is intended to expose qualitative differences between moderate and wide
% deception. The mutation rates $\{0.5/n, 1/n, 2/n, 5/n\}$ are shared with
% suites 05 and 06 to allow comparison of regret shape across barrier types at
% equal rates.
%
% Profile analysis is disabled.

Suite 04 evaluates Trap with four $(1+1)$-\acrshort{ea} mutation rates
($0.5/n$, $1/n$, $2/n$, $5/n$) and SA-Log-d\_4.0 across four deception widths
($k \in \{5,8,10,20\}$, $n=100$). The EA rates are fixed across all problems;
SA uses $d=4.0$ by default with a single override to $d=8.0$ on Trap-k20.
Trap-k1 is excluded because all algorithms are uniformly driven to the deceptive
attractor, producing uninformative flat results. The chosen EA rates are shared
with suites 05 and 06 to support cross-suite comparison of regret shape across
barrier types. Profile analysis is disabled.

\subsection{Suite 05: Jump Function Mutation Rate}

% Suite 05 studies five Jump instances ($k \in \{2, 3, 5, 7, 10\}$, $n=100$)
% under three $(1+1)$-\acrshort{ea} variants. Two variants use fixed
% rates: $1/n$ (0.01)
% and $2/n$ (0.02), uniform across all problems. The third uses a
% problem-adaptive rate $k/n$, set via per-problem overrides to
% $\{0.02, 0.03, 0.05, 0.07, 0.10\}$ for $k \in \{2, 3, 5, 7, 10\}$
% respectively; the \texttt{defaults} block carries 0.03 as a fallback that is
% overridden for every problem actually in the suite.

% The $0.5/n$ rate present in suite 04 is absent here. At rate $0.5/n$ the
% expected number of bit-flips per step is half a bit, which is strictly worse
% than rate $1/n$ for crossing a $k$-bit gap: lower mutation rate never helps
% cross a gap that requires $k$ simultaneous flips, so the curve adds no
% insight beyond the $1/n$ baseline.

% The budget list extends to 20\,000 and includes six points:\\
% $\{500, 1000, 2000, 5000, 10000, 20000\}$. Cumulative regret is plotted on a
% linear $y$-axis (\texttt{log\_y: false}), while instantaneous regret uses
% \texttt{log\_y: true}. This distinction matters: during the plateau phase
% (when the algorithm is trapped at $\text{ones}=n-k$), cumulative regret grows
% linearly with $t$, and this linear slope is the signal of interest; a
% logarithmic $y$-axis would compress it to near-zero and make it invisible.

% Profile analysis is disabled; the related profile treatment of Jump is
% deferred to suite 09.
%
Suite 05 studies five Jump instances ($k \in \{2,3,5,7,10\}$, $n=100$) under
three $(1+1)$-\acrshort{ea} variants: fixed rates $1/n$ and $2/n$, and a
problem-adaptive rate $k/n$ set via per-problem overrides. The lower rate
$0.5/n$ is omitted because it offers no advantage over $1/n$ for crossing a
$k$-bit gap. Budgets range from 500 to 20\,000. Instantaneous regret is plotted
on a log scale, while cumulative regret uses a linear scale so that the
trapped-phase slope remains visible. Profile analysis is disabled; the
longer-horizon Jump study is deferred to suite 09.

\subsection{Suite 06: Plateau Function Mutation Rate}

% Suite 06 mirrors suite 05 on four Plateau instances ($k \in \{2, 5, 10, 20\}$,
% $n=100$) to enable direct trajectory comparison between Jump and Plateau at
% structurally equivalent barrier widths. The algorithm set is three
% $(1+1)$-\acrshort{ea}
% variants at rates $1/n$, $2/n$, and $5/n$, plus SA-Log-d\_4.0 with no
% per-problem overrides.

% Two differences from suite 05 are worth noting. First, the $k/n$
% problem-adaptive \acrshort{ea} variant is replaced by the $5/n$
% fixed-rate variant. The
% adaptive rate had a clear theoretical motivation for Jump (it maximizes the
% probability of exactly $k$ simultaneous flips needed to cross the gap) but
% has no equivalent motivation for Plateau, where the neutral walk does not
% require a specific number of simultaneous flips. Second, SA is added as a
% third algorithm class to assess whether temperature-driven acceptance provides
% any benefit on neutral landscapes; there is no analogous SA entry in suite 05.

% The Plateau $k$ values ($\{2, 5, 10, 20\}$) do not exactly mirror the Jump $k$
% values ($\{2, 3, 5, 7, 10\}$) from suite 05. The two suites share $k=2$ and
% $k=5$, enabling a direct Plateau--Jump comparison at those widths. The
% remaining Plateau values ($k=10$, $k=20$) push into regimes where convergence
% within the 20\,000-evaluation budget is unlikely for all algorithms, providing
% longer plateau durations in the trajectory plots.

% Profile analysis is disabled.
% 
Suite 06 mirrors suite 05 on Plateau instances ($k \in \{2,5,10,20\}$, $n=100$)
to compare cumulative regret trajectory shape across barrier types at
structurally equivalent widths. The algorithm set is three
$(1+1)$-\acrshort{ea} variants at rates $1/n$, $2/n$, and $5/n$, plus
SA-Log-d\_4.0.

The adaptive $k/n$ rate from suite 05 is omitted: it was motivated by the need
to maximize the probability of exactly $k$ simultaneous flips for the Jump gap,
a rationale that does not apply to the neutral walk on Plateau. SA is added to
assess whether temperature-driven acceptance confers any advantage on
a landscape
where the neutral zone carries no fitness penalty. The shared values $k=2$ and
$k=5$ support direct comparison with suite 05. Profile analysis is disabled.

\subsection{Suite 07: Hierarchical IFF Scaling}

% Suite 07 evaluates three HIFF instances ($n \in \{32, 64, 128\}$) under six
% algorithms: \acrshort{rls}, four $(1+1)$-\acrshort{ea} variants
% with per-problem mutation
% rates, and SA-Log-d\_4.0.

% The EA mutation rates are set via \texttt{by\_problem} overrides to exact
% decimal values corresponding to $\{0.5/n, 1/n, 2/n, 3/n\}$ at each of the
% three problem sizes:

% \begin{table}[htbp]
% \centering
% \caption{Per-problem mutation rates in suite 07. Values correspond to exact
%          fractions of $1/n$ for each $n$.}
% \label{tab:suite07-rates}
% \small
% \begin{tabular}{lrrr}
% \toprule
% Algorithm & $n=32$ & $n=64$ & $n=128$ \\
% \midrule
% (1+1)-EA-mu\_0.5-by-n & $\nicefrac{1}{64} \approx 0.01563$ &
% $\nicefrac{1}{128} \approx 0.00781$ & $\nicefrac{1}{256} \approx 0.00391$ \\
% (1+1)-EA-mu\_1-by-n   & $\nicefrac{1}{32} \approx 0.03125$ &
% $\nicefrac{1}{64}  \approx 0.01563$ & $\nicefrac{1}{128} \approx 0.00781$ \\
% (1+1)-EA-mu\_2-by-n   & $\nicefrac{2}{32} \approx 0.06250$ &
% $\nicefrac{2}{64}  \approx 0.03125$ & $\nicefrac{2}{128} \approx 0.01563$ \\
% (1+1)-EA-mu\_3-by-n   & $\nicefrac{3}{32} \approx 0.09375$ &
% $\nicefrac{3}{64}  \approx 0.04688$ & $\nicefrac{3}{128} \approx 0.02344$ \\
% \bottomrule
% \end{tabular}
% \end{table}

% SA-Log-d\_4.0 uses $d=4.0$ uniformly with no per-problem overrides; it is not
% tuned to any specific $n$ value. The \texttt{budget\_for\_plots} for $n=128$
% is 50\,000, which coincides with the largest available budget. There is no
% expectation that algorithms converge at $n=128$ within this budget; the value
% is set to show the full trajectory for a problem that is expected to remain
% unsolved, providing an informative contrast with $n=32$ and $n=64$ in the
% same comparison.

% Profile analysis is disabled.

Suite 07 evaluates \acrshort{hiff} at $n \in \{32,64,128\}$ under
\acrshort{rls}, four $(1+1)$-\acrshort{ea} variants at rates
$\{0.5/n,1/n,2/n,3/n\}$ (exact per-problem decimal values listed in
Table~\ref{tab:suite07-rates}), and SA-Log-d\_4.0 at a fixed $d=4.0$.
\acrshort{rls} is expected to stagnate at the first hierarchical level for
$n \geq 64$ since a single bit-flip disrupts any uniform sub-block, anchoring
the comparison. For $n=128$, \texttt{budget\_for\_plots} is set to the maximum
available (50\,000); convergence is not expected, and the unsolved trajectory
provides an informative contrast to the partially solved $n=32$ and $n=64$
cases. Profile analysis is disabled.

\begin{table}[H]
  \centering
  \caption{Per-problem mutation rates in suite 07. Values are set via
    \texttt{by\_problem} overrides to exact decimal representations of
  $\{0.5/n,1/n,2/n,3/n\}$ at each problem size.}
  \label{tab:suite07-rates}
  \small
  \begin{tabular}{@{}lrrr@{}}
    \toprule
    \textbf{Algorithm} & $n=32$ & $n=64$ & $n=128$ \\
    \midrule
    EA-mu\_0.5-by-n
    & $\nicefrac{1}{64}\approx0.01563$
    & $\nicefrac{1}{128}\approx0.00781$
    & $\nicefrac{1}{256}\approx0.00391$ \\
    EA-mu\_1-by-n
    & $\nicefrac{1}{32}\approx0.03125$
    & $\nicefrac{1}{64}\approx0.01563$
    & $\nicefrac{1}{128}\approx0.00781$ \\
    EA-mu\_2-by-n
    & $\nicefrac{2}{32}\approx0.06250$
    & $\nicefrac{2}{64}\approx0.03125$
    & $\nicefrac{2}{128}\approx0.01563$ \\
    EA-mu\_3-by-n
    & $\nicefrac{3}{32}\approx0.09375$
    & $\nicefrac{3}{64}\approx0.04688$
    & $\nicefrac{3}{128}\approx0.02344$ \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Suite 09: Jump Plateau Duration}

Suite 09 uses the same five Jump instances and two EA variants as suite 05
(rates $1/n$ and $k/n$ with identical per-problem overrides), extending the
budget list to $\{1000, 5000, 10000, 25000, 50000\}$ and raising
\texttt{budget\_for\_plots} to 50\,000 for the larger $k$ values. Where suite 05
captures plateau onset for small $k$, suite 09 captures plateau duration across
all $k$ values and, for small $k$ at the adaptive rate, potentially the plateau
exit. Profile analysis is disabled.

\subsection{Suite 10: Barrier Structure Comparison}

% Suite 10 compares three barrier archetypes at equal nominal width $k=5$
% ($n=100$): Jump-k5, Trap-k5, and Plateau-k5. Four $(1+1)$-EA variants at
% rates $0.5/n$, $1/n$, $2/n$, and $5/n$ are applied with no per-problem
% overrides, meaning the same rate is used on all three barrier types. The
% budget list extends to 50\,000 to allow the $k=5$ barrier to be crossed at
% moderate rates within the observation window.

% The motivation for fixing $k=5$ and varying the barrier type (rather than
% fixing the type and varying $k$) is that Jump-k5, Trap-k5, and Plateau-k5
% all have $\mathbb{E}[\text{\acrshort{ttfo}}] = \Theta(n^{k})$ at
% rate $1/n$, so \acrshort{ttfo}
% distributions cannot separate them when $P(\text{opt}) \approx 0$.
% Cumulative regret trajectories, by contrast, accumulate differently depending
% on the fitness gradient before and during the barrier: Trap accumulates regret
% from the first step (the gradient points away from the optimum throughout),
% Jump accumulates low regret during hill-climbing and then high regret during
% the trapped phase, and Plateau behaves similarly to Jump but without a fitness
% penalty in the neutral zone. Whether this theoretical distinction is visible
% in the trajectory plots at $n=100$ and $k=5$ is what the suite is designed to
% test.

% Pairwise annotations use \texttt{(1+1)-\acrshort{ea}-mu\_1-by-n} as
% the reference
% algorithm on box-plots, performance profiles, and \acrshort{ttfo}
% distributions.
% Profile analysis is disabled.
%
Suite 10 compares Jump-k5, Trap-k5, and Plateau-k5 at $n=100$ under four
$(1+1)$-\acrshort{ea} mutation rates: $0.5/n$, $1/n$, $2/n$, and $5/n$, applied
uniformly across all three barrier types with budgets to 50\,000. Since all
three barriers share $\mathbb{E}[\text{\acrshort{ttfo}}] = \Theta(n^{k})$ at
rate $1/n$, \acrshort{ttfo} distributions cannot separate them when
$P(\text{opt}) \approx 0$; cumulative regret trajectories accumulate
differently depending on the fitness gradient before and during the barrier, and
testing whether this distinction is visible at $n=100$, $k=5$ is the suite's
primary purpose. Pairwise annotations use
\texttt{(1+1)-\acrshort{ea}-mu\_1-by-n} as the reference. Profile analysis is
disabled.

\subsection{Suite 11: SA Cooling Schedule Diagnostics}

% Suite 11 evaluates eight SA variants and one hill-climber baseline on
% OneMax and LeadingOnes ($n=100$). The eight SA algorithms are: six
% logarithmic cooling variants with $d \in \{0.5, 1.0, 2.0, 4.0, 8.0,
% 16.0\}$, one linear cooling instance (SA-Lin), and one exponential
% cooling instance (SA-Exp). All SA entries carry empty
% \texttt{by\_problem} blocks; the same schedule applies to both
% problems. \acrshort{rls} is the ninth algorithm, serving as a
% zero-temperature reference.

% The linear and exponential schedules use the default parameter values from
% the implementation: linear cooling uses $T_0=1.0$, $T_f=0.01$,
% $\text{max\_iter}=10000$; exponential cooling uses $T_0=1.0$, $\alpha=0.95$.
% These parameters are not configurable via \texttt{params} keys in this suite's
% YAML (the \texttt{params} sub-blocks are empty) so the schedule behavior
% follows the implementation defaults. Both linear and exponential schedules
% reach their minimum temperature within the available budget (linear at
% $t=10000$, exponential asymptotically but effectively by $t \approx 90$
% at $\alpha=0.95$), after which SA reduces to a near-deterministic
% hill-climber. This is in contrast to the logarithmic schedules, which remain
% strictly positive for all finite $t$.

% Pairwise annotations use \acrshort{rls} as the reference on box-plots and
% performance profiles. The selection of \acrshort{rls} as the reference (rather
% than the best-performing SA variant) means that any \acrshort{sa}
% configuration which
% underperforms \acrshort{rls} is directly identifiable in the annotated plots.
% Profile analysis is disabled.

Suite 11 evaluates six logarithmic SA schedules ($d \in
\{0.5,1.0,2.0,4.0,8.0,16.0\}$), one linear schedule (SA-Lin), one exponential
schedule (SA-Exp), and \acrshort{rls} as a zero-temperature reference on OneMax
and LeadingOnes ($n=100$). The linear and exponential variants use
implementation defaults (linear: $T_0=1.0$, $T_f=0.01$,
$\text{max\_iter}=10\,000$; exponential: $T_0=1.0$, $\alpha=0.95$); both reach
their minimum temperature within the budget (exponential effectively by
$t \approx 90$), after which SA reduces to a near-deterministic hill-climber.
This structural distinction between collapsing and sustaining schedules is the
primary hypothesis the suite is designed to expose via instantaneous regret
trajectories. Pairwise annotations use \acrshort{rls} as the reference so that
any SA configuration underperforming plain hill-climbing is directly
identifiable. Profile analysis is disabled.

\subsection{Suite 12: NK Saturation}

Suite 12 examines five NK instances ($k \in \{0,2,4,6,8\}$, $n=20$, seed 42)
under \acrshort{rls}, the $(1+1)$-\acrshort{ea}, and SA-Log-d\_4.0 with
per-$k$ cooling parameter escalation (values in
Table~\ref{tab:suite12-d}). The landscape realizations at
$k \in \{2,4,6,8\}$ are identical to suite 03, enabling direct comparison. The
addition of $k=0$ (fully separable) provides a saturation baseline. Unlike
suite 03, SA is calibrated per problem rather than forming a grid; suite 12
therefore asks whether adequate temperature maintenance changes the saturation
outcome rather than how the cooling--epistasis interaction behaves across a full
parameter sweep. Profile analysis is enabled; the
\texttt{cr\_profile\_verification} plot is omitted because the NK fitness
function is continuous and the integer-grid profile identity does not apply
exactly.

\begin{table}[H]
  \centering
  \caption{SA-Log cooling parameter $d$ per NK instance in suite 12.
    Values increase monotonically with epistasis $k$ to compensate for
  deeper local optima.}
  \label{tab:suite12-d}
  \small
  \begin{tabular}{@{}lrrrrr@{}}
    \toprule
    \textbf{NK instance} & NK-k0 & NK-k2 & NK-k4 & NK-k6 & NK-k8 \\
    \midrule
    $d$ & 4.0 & 4.0 & 8.0 & 8.0 & 12.0 \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Suite 13: Regret Scaling with Problem Size}

Suite 13 runs the $(1+1)$-\acrshort{ea} at rate $1/n$ on four OneMax instances
($n \in \{25,50,100,200\}$) over 60 runs (the second-highest run count in the
study) to produce empirical $\mathbb{E}[\text{CR}(T)]$ estimates suitable for
fitting a scaling exponent. Using a single algorithm isolates the scaling
behavior of the regret metric from algorithm-to-algorithm differences.
\texttt{budget\_for\_plots} values are set somewhat above $5n\log_2 n$ so that
the full post-convergence plateau is visible in trajectory plots. Profile
analysis is disabled.

\subsection{Suite 14: Population Dynamics}

Suite 14 compares the $(1+1)$-\acrshort{ea} against four
$(\mu+\lambda)$-\acrshort{ea} variants ($(1+1)$ via framework, $(5+5)$,
$(10+20)$, and $(20+10)$) on OneMax, LeadingOnes, TwoMax, and Jump-3
($n=100$). All entries use mutation rate $1/n$ with no per-problem overrides,
isolating the effect of population structure. Including both the standalone
$(1+1)$-\acrshort{ea} and its framework equivalent ($\mu=1$, $\lambda=1$) acts
as an implementation consistency check: any divergence between them indicates
framework overhead rather than a population effect. Profile analysis is
disabled.

\section{Cross-Suite Design Relationships}

Several structural relationships between suites affect which results can be
meaningfully compared. Table~\ref{tab:cross-suite} summarizes the three
principal links.

\begin{table}[H]
  \centering
  \caption{Principal cross-suite design relationships.}
  \label{tab:cross-suite}
  \small
  \begin{tabular}{@{}p{2.5cm}p{2.6cm}p{4.0cm}p{3.5cm}@{}}
    \toprule
    \textbf{Relationship} & \textbf{Suites} & \textbf{Shared} &
    \textbf{Differs} \\
    \midrule
    Barrier types at equal rates
    & 04, 05, 06, 10
    & EA rates $1/n$ and $2/n$; problems at same nominal $k$
    & Barrier type (Trap/Jump/Plateau); additional rates per suite \\
    \addlinespace
    Jump at two time horizons
    & 05, 09
    & All 5 Jump instances; EA variants at $1/n$ and $k/n$
    & Budget list and \texttt{budget\_for\_plots} (onset vs.\ duration) \\
    \addlinespace
    NK at two SA designs
    & 03, 12
    & $n=20$, seed 42; shared $k\in\{2,4,6,8\}$; identical landscapes
    & Suite 03: $6\times5$ $d$-grid, no tuning; Suite 12: per-$k$ $d$
    escalation, adds $k=0$ \\
    \bottomrule
  \end{tabular}
\end{table}

Suites 04, 05, and 06 share the fixed rates $1/n$ and $2/n$, making Trap-$k$,
Jump-$k$, and Plateau-$k$ results directly comparable at equal $k$; suite 10
consolidates this at $k=5$ across all four EA rates simultaneously. Suites 05
and 09 are identical in problem and algorithm design, differing only in budget
horizon: suite 05 captures plateau onset, suite 09 captures plateau duration.
Suites 03 and 12 share landscape realizations at $k \in \{2,4,6,8\}$ but
differ in SA design: suite 03 provides the clean grid for studying the
cooling--epistasis interaction, while suite 12 provides the per-$k$ calibrated
view of saturation.
